---
title: "Chromosome-scale feature density plots"
author: 
 - "Michael Mansfield"
 - "Charles Plessy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Chromosome-scale feature density plots}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(cache = TRUE)
```

# Introduction

In this vignette, we develop some of the materials needed to produce plots that summarize genomic data at chromosome-scale.

The core functions used here are maintained in our _GenomicBreaks_ R package,
which is fully documented at: <https://oist.github.io/GenomicBreaks>.

## Load packages

# Load R packages and data

```{r load_packages_and_data}
library('OikScrambling') |> suppressPackageStartupMessages()
library('viridis') |> suppressPackageStartupMessages()
library('dplyr') |> suppressPackageStartupMessages()
genomes <- OikScrambling:::loadAllGenomes()
transcripts <- OikScrambling:::loadAllTranscriptsGR()
reps <- OikScrambling:::loadAllRepeats()
load("BreakPoints.Rdata")
```

### Creating chromosome-scale plots
In this vignette, we create the plumbing needed to generate summary plots that show windows of genomic features on a chromosome scale. 

#### Summarizing genomic windows of annotated `GRanges` objects
Below, I create some of the functions needed to produce these plots.

```{r chromosome_plot_helper_fxns}
# Make a BSgenome into a GRanges object, one range per chromosome
BSgenomeToGR <- function(genome) {
  as(seqinfo(genome), "GRanges")
}
(tmp <- genomes$Oki |> BSgenomeToGR())

# Make a genome into a bunch of windows of length N
genomeGRToWindows <- function(gr, windowSize=10000) {
  tile(gr, w=windowSize) |> unlist() |> unname()
}
(tmp <- genomes$Oki |> BSgenomeToGR() |> genomeGRToWindows())

# Match ranges to a single bin by reducing each range to its center.
# Returns the index of the matched bins
matchToOneBin <- function(ranges, bins) {
  # Ensure that a range overlaps with only one bin
  safeRanges <- ranges |> resize(1, fix="center")
  o <- findOverlaps(safeRanges, bins, type = 'within')
  subjectHits(o)
}
(matchToOneBin(transcripts$Oki, tmp)) |> head()

# Apply one function to data grouped by bins, and array it according to the
# original bins (that is, insert NAs for bins that contained no data).
# This function only works as intended if the bins fully tile the genome.
binApply <- function(f, bins, data, index, ...) {
  res <- rep(NA, length(bins))
  vals <-tapply(data, index, f, ...)
  res[as.numeric(names(vals))] <- unname(vals)
  res
}
(binApply(mean, tmp, transcripts$Oki$dNdS_GUIDANCE2, matchToOneBin(transcripts$Oki, tmp), na.rm = TRUE)) |> head(100)

# Intersect each genome window with some other GRange (meta, as in metadata)
metaPerWindow <- function(genomeWindows, meta, meta_prefix=NULL) {
  idx <- matchToOneBin(meta, genomeWindows)
  for (name in colnames(mcols(meta))) {
    data <- mcols(meta)[[name]]
    if(is.numeric(data)) {
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".count")]]       <- binApply(length, genomeWindows, data, idx)
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".median")]]      <- binApply(median, genomeWindows, data, idx, na.rm = TRUE)
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".mean")]]        <- binApply(mean,   genomeWindows, data, idx, na.rm = TRUE)
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".sd")]]          <- binApply(sd,     genomeWindows, data, idx, na.rm = TRUE)
    } else if(is.character(data) ) {
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".count")]]       <- binApply(length, genomeWindows, data, idx)
    } else if(is.factor(data)) {
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".count.total")]] <- binApply(length, genomeWindows, data, idx)
      if(length(levels(data)) > 1) {
        binned_list_of_tables <- binApply(table,  genomeWindows, data, idx)
        binned_DataFrame <- do.call(rbind, binned_list_of_tables) |> DataFrame()
        colnames(binned_DataFrame) <- paste0(meta_prefix, name, ".", colnames(binned_DataFrame), ".count.total")
        mcols(genomeWindows) <- cbind(mcols(genomeWindows), binned_DataFrame)
      }
    }
  }
  genomeWindows
}
(metaPerWindow(genomeWindows=tmp, transcripts$Oki, meta_prefix = 'transcripts.'))

# Function to do everything
grWindows <- function(genome, meta, windowSize=10000, meta_prefix=NULL){
  genome <- BSgenomeToGR(genome)                                 # Make GRanges from genome
  genome <- genomeGRToWindows(genome, windowSize=windowSize)     # Make a windowed GRange from chromosome GRanges
  mpw <-    metaPerWindow(genome, meta, meta_prefix=meta_prefix) # Make metadata column windows and intersect
  mpw
}
```

## Chromosome plots
### Transcripts
In the following sections we annotate and initialize the `chrw` object, which is a `SimpleList()` containing windowed chromosome regions. 

We use these windows to evaluate the distribution of various features across the chromosomes. This section concerns one-dimensional features.

```{r chromosome_plots_transcripts}
windowSize <- 2e5
bins_Oki <- genomes$Oki |> BSgenomeToGR() |> genomeGRToWindows(windowSize)
bins_Osa <- genomes$Osa |> BSgenomeToGR() |> genomeGRToWindows(windowSize)
bins_Bar <- genomes$Bar |> BSgenomeToGR() |> genomeGRToWindows(windowSize)

chrw <- SimpleList()
chrw$Oki <- grWindows(genome=genomes$Oki, meta=transcripts$Oki, windowSize=windowSize, meta_prefix = 'transcripts.')
chrw$Oki <- flagLongShort(chrw$Oki, longShort$OKI2018.I69)
chrw$Osa <- grWindows(genome=genomes$Osa, meta=transcripts$Osa, windowSize=windowSize, meta_prefix = 'transcripts.')
chrw$Osa <- flagLongShort(chrw$Osa, longShort$OSKA2016v1.9)
chrw$Bar <- grWindows(genome=genomes$Bar, meta=transcripts$Bar, windowSize=windowSize, meta_prefix = 'transcripts.')
chrw$Bar <- flagLongShort(chrw$Bar, longShort$Bar2.p4)
```

```{r chromosome_plots_transcripts_oki, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Density of transcript names (i.e., gene counts).
ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.tx_name.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth() + ggtitle(paste("Okinawa: transcripts per window (size = ", windowSize/1000, " kbp)", sep="")) +
  geom_vline( data = (longShort$OKI2018.I69 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame()
            , aes(xintercept = start), col = "gray")

ggplot(as.data.frame(chrw$Oki)  |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median) + geom_point() + facet_wrap(seqnames ~ Arm) + ggtitle(paste("Okinawa: dN/dS per window (size = ", windowSize/1000, " kbp)", sep=""))

ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: dN/dS per window (size = ", windowSize/1000, " kbp)", sep=""))
```


```{r chromosome_plots_transcripts_osa, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Density of transcript names (i.e., gene counts).
ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.tx_name.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth() + ggtitle(paste("Osaka: transcripts per window (size = ", windowSize/1000, " kbp)", sep=""))

ggplot(as.data.frame(chrw$Osa)  |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median) + geom_point() + facet_wrap(seqnames ~ Arm) + ggtitle(paste("Osaka: dN/dS per window (size = ", windowSize/1000, " kbp)", sep=""))

ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median) + geom_point() + facet_wrap(~seqnames) + geom_smooth() + ggtitle(paste("Osaka: dN/dS per window (size = ", windowSize/1000, " kbp)", sep="")) 
```

```{r chromosome_plots_transcripts_bar, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Density of transcript names (i.e., gene counts).
ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.tx_name.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth() + ggtitle(paste("Barcelona: transcripts per window (size = ", windowSize/1000, " kbp)", sep=""))

ggplot(as.data.frame(chrw$Bar)  |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median) + geom_point() + facet_wrap(seqnames ~ Arm) + ggtitle(paste("Barcelona: dN/dS per window (size = ", windowSize/1000, " kbp)", sep=""))

ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median) + geom_point() + facet_wrap(~seqnames) + geom_smooth() + ggtitle(paste("Barcelona: dN/dS per window (size = ", windowSize/1000, " kbp)", sep=""))
```

#### Transcripts: joined gene density plot
```{r chromosome_plots_transcripts_joined, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Okinawa
p1 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_id.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: genes per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('transcripts')

# Osaka
p2 <- ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_id.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Osaka: genes per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('transcripts')

# Barcelona
p3 <- ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_id.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Barcelona: genes per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('transcripts')

ggpubr::ggarrange(p1, p2, p3, ncol=1)
```

#### Transcripts: joined dN/dS plot
```{r chromosome_plots_transcripts_dnds_joined, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: dN/dS per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab("dN/dS")

p2 <- ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Osaka: dN/dS per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab("dN/dS")

p3 <- ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Barcelona: dN/dS per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab("dN/dS")

ggpubr::ggarrange(p1, p2, p3, ncol=1)
```


### Repeats
In this section we add repeat data to the `chrw` object. Since the windows are the same for every genome, this is simple accomplished with a `cbind` operation to the windowed object's `mcols`.

```{r chromosome_plots_repeats, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Since the coordinates of the windows are the same, you can simply cbind more metadata columns to the data frame.
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=reps$Oki, windowSize=windowSize, meta_prefix = 'repeats.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=reps$Osa, windowSize=windowSize, meta_prefix = 'repeats.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=reps$Bar, windowSize=windowSize, meta_prefix = 'repeats.')))

```

```{r chromosome_plots_repeats_1, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=repeats.Class.count.total) + geom_point() + facet_wrap(seqnames~Arm) + geom_smooth() + scale_y_log10()

ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=repeats.Class.count.total) + geom_point() + facet_wrap(seqnames~Arm) + geom_smooth() + scale_y_log10()

ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=repeats.Class.count.total) + geom_point() + facet_wrap(seqnames~Arm) + geom_smooth() + scale_y_log10()
```

We can use the `[mcol].counts.tables` metadata columns to calculate statistics between different genomic regions.

```{r chromosome_plots_repeats_2_oki, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.ltr.1.count.total", "repeats.Class.rnd.count.total", "repeats.Class.SINE.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.ltr.1.count.total", "repeats.Class.rnd.count.total", "repeats.Class.SINE.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Oki: distribution of repeat classes across genome")

```


```{r chromosome_plots_repeats_2_osa, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.rnd.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.rnd.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Osa: distribution of repeat classes across genome")

```



```{r chromosome_plots_repeats_2_bar, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.rnd.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.rnd.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Bar: distribution of repeat classes across genome")

```

```{r chromosome_plots_repeats_joined, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=repeats.type.count.total, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: repeats per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('repeats')

p2 <- ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=repeats.type.count.total, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Osa: repeats per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('repeats')

p3 <- ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=repeats.type.count.total, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Barcelona: repeats per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('repeats')

ggpubr::ggarrange(p1, p2, p3, ncol=1)
```

## `GenomicBreaks` objects: Scores and widths

In the next sections, we explore the distribution of alignment parameters (scores and widths) for different populations vs. chromosomal position.

```{r chromosome_plots_scores, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Okinawa
chrw$Oki$Oki_Osa.score.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Osa), matchToOneBin(gbs$Oki_Osa, bins_Oki))
chrw$Oki$Oki_Aom.score.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Aom), matchToOneBin(gbs$Oki_Aom, bins_Oki))
chrw$Oki$Oki_Bar.score.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Bar), matchToOneBin(gbs$Oki_Bar, bins_Oki))
chrw$Oki$Oki_Nor.score.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Nor), matchToOneBin(gbs$Oki_Nor, bins_Oki))

# Osaka
chrw$Osa$Osa_Oki.score.mean <- binApply(mean, na.rm = T, bins_Osa, score(gbs$Osa_Oki), matchToOneBin(gbs$Osa_Oki, bins_Osa))
chrw$Osa$Osa_Kum.score.mean <- binApply(mean, na.rm = T, bins_Osa, score(gbs$Osa_Kum), matchToOneBin(gbs$Osa_Kum, bins_Osa))
chrw$Osa$Osa_Bar.score.mean <- binApply(mean, na.rm = T, bins_Osa, score(gbs$Osa_Bar), matchToOneBin(gbs$Osa_Bar, bins_Osa))
chrw$Osa$Osa_Nor.score.mean <- binApply(mean, na.rm = T, bins_Osa, score(gbs$Osa_Nor), matchToOneBin(gbs$Osa_Nor, bins_Osa))

# Barcelona
chrw$Bar$Bar_Osa.score.mean <- binApply(mean, na.rm = T, bins_Bar, score(gbs$Bar_Osa), matchToOneBin(gbs$Bar_Osa, bins_Bar))
chrw$Bar$Bar_Aom.score.mean <- binApply(mean, na.rm = T, bins_Bar, score(gbs$Bar_Aom), matchToOneBin(gbs$Bar_Aom, bins_Bar))
chrw$Bar$Bar_Oki.score.mean <- binApply(mean, na.rm = T, bins_Bar, score(gbs$Bar_Oki), matchToOneBin(gbs$Bar_Oki, bins_Bar))
chrw$Bar$Bar_Kum.score.mean <- binApply(mean, na.rm = T, bins_Bar, score(gbs$Bar_Kum), matchToOneBin(gbs$Bar_Kum, bins_Bar))
```

```{r chromosome_plots_scores_oki, dependson="chromosome_plots_scores", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Osa.score.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth() +
  scale_y_log10() +
  ggtitle("Oki-Osa: Chromosomal position vs. alignment score")

ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Bar.score.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth() +
  scale_y_log10() + 
  ggtitle("Oki-Bar: Chromosomal position vs. alignment score")


chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki_Osa.score.mean", "Oki_Aom.score.mean", "Oki_Bar.score.mean", "Oki_Nor.score.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Oki_Osa.score.mean", "Oki_Aom.score.mean", "Oki_Bar.score.mean", "Oki_Nor.score.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. alignment score")

```


```{r chromosome_plots_scores_joined, dependson="chromosome_plots_scores", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki_Osa.score.mean", "Oki_Aom.score.mean", "Oki_Bar.score.mean", "Oki_Nor.score.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Oki_Osa.score.mean", "Oki_Aom.score.mean", "Oki_Bar.score.mean", "Oki_Nor.score.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. alignment score")

p2 <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Osa_Oki.score.mean", "Osa_Kum.score.mean", "Osa_Bar.score.mean", "Osa_Nor.score.mean" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Osa_Oki.score.mean", "Osa_Kum.score.mean", "Osa_Bar.score.mean", "Osa_Nor.score.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Osaka: Chromosomal position vs. alignment score")

p3 <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Bar_Oki.score.mean", "Bar_Kum.score.mean", "Bar_Osa.score.mean", "Bar_Aom.score.mean" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Bar_Oki.score.mean", "Bar_Kum.score.mean", "Bar_Osa.score.mean", "Bar_Aom.score.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Barcelona: Chromosomal position vs. alignment score")

ggpubr::ggarrange(p1, p2, p3, ncol=1)
```


In the following section I divide the scores by the widths of the alignments, creating a scaled score.

```{r chromosome_plots_scores_norm, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
chrw$Oki$Oki_Osa.score.norm.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Osa)/lengths(gbs$Oki_Osa), matchToOneBin(gbs$Oki_Osa, bins_Oki))
chrw$Oki$Oki_Aom.score.norm.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Aom)/lengths(gbs$Oki_Aom), matchToOneBin(gbs$Oki_Aom, bins_Oki))
chrw$Oki$Oki_Bar.score.norm.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Bar)/lengths(gbs$Oki_Bar), matchToOneBin(gbs$Oki_Bar, bins_Oki))
chrw$Oki$Oki_Nor.score.norm.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Nor)/lengths(gbs$Oki_Nor), matchToOneBin(gbs$Oki_Nor, bins_Oki))

ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Osa.score.norm.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth() +
  scale_y_log10() +
  ggtitle("Oki-Osa: Chromosomal position vs. scaled alignment score")


ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Bar.score.norm.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth() +
  scale_y_log10() +
  ggtitle("Oki-Bar: Chromosomal position vs. scaled alignment score")


chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki_Osa.score.norm.mean", "Oki_Aom.score.norm.mean", "Oki_Bar.score.norm.mean", "Oki_Nor.score.norm.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Oki_Osa.score.norm.mean", "Oki_Aom.score.norm.mean", "Oki_Bar.score.norm.mean", "Oki_Nor.score.norm.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. scaled alignment score")

```

## Alignment widths

```{r chromosome_plots_widths, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Okinawa
chrw$Oki$Oki_Osa.width.mean <- binApply(mean, na.rm = T, bins_Oki, width(gbs$Oki_Osa), matchToOneBin(gbs$Oki_Osa, bins_Oki))
chrw$Oki$Oki_Aom.width.mean <- binApply(mean, na.rm = T, bins_Oki, width(gbs$Oki_Aom), matchToOneBin(gbs$Oki_Aom, bins_Oki))
chrw$Oki$Oki_Bar.width.mean <- binApply(mean, na.rm = T, bins_Oki, width(gbs$Oki_Bar), matchToOneBin(gbs$Oki_Bar, bins_Oki))
chrw$Oki$Oki_Nor.width.mean <- binApply(mean, na.rm = T, bins_Oki, width(gbs$Oki_Nor), matchToOneBin(gbs$Oki_Nor, bins_Oki))


# Osaka
chrw$Osa$Osa_Oki.width.mean <- binApply(mean, na.rm = T, bins_Osa, width(gbs$Osa_Oki), matchToOneBin(gbs$Osa_Oki, bins_Osa))
chrw$Osa$Osa_Kum.width.mean <- binApply(mean, na.rm = T, bins_Osa, width(gbs$Osa_Kum), matchToOneBin(gbs$Osa_Kum, bins_Osa))
chrw$Osa$Osa_Bar.width.mean <- binApply(mean, na.rm = T, bins_Osa, width(gbs$Osa_Bar), matchToOneBin(gbs$Osa_Bar, bins_Osa))
chrw$Osa$Osa_Nor.width.mean <- binApply(mean, na.rm = T, bins_Osa, width(gbs$Osa_Nor), matchToOneBin(gbs$Osa_Nor, bins_Osa))

# Barcelona
chrw$Bar$Bar_Osa.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Osa), matchToOneBin(gbs$Bar_Osa, bins_Bar))
chrw$Bar$Bar_Aom.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Aom), matchToOneBin(gbs$Bar_Aom, bins_Bar))
chrw$Bar$Bar_Oki.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Oki), matchToOneBin(gbs$Bar_Oki, bins_Bar))
chrw$Bar$Bar_Kum.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Kum), matchToOneBin(gbs$Bar_Kum, bins_Bar))
```

```{r chromosome_plots_widths_oki, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Osa.width.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth() +
  scale_y_log10() +
  ggtitle("Oki-Osa: Chromosomal position vs. aligned segment width")

ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Osa.width.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth() +
  scale_y_log10() +
  ggtitle("Oki-Bar: Chromosomal position vs. aligned segment width")

chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki_Osa.width.mean", "Oki_Aom.width.mean", "Oki_Bar.width.mean", "Oki_Nor.width.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Oki_Osa.width.mean", "Oki_Aom.width.mean", "Oki_Bar.width.mean", "Oki_Nor.width.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. aligned segment width")
```

```{r chromosome_plots_widths_joined, dependson="chromosome_plots_widths", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki_Osa.width.mean", "Oki_Aom.width.mean", "Oki_Bar.width.mean", "Oki_Nor.width.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Oki_Osa.width.mean", "Oki_Aom.width.mean", "Oki_Bar.width.mean", "Oki_Nor.width.mean")) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(span = 0.25, se = FALSE) +
  ggtitle("Okinawa: Chromosomal position vs. aligned segment width")

p2 <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Osa_Oki.width.mean", "Osa_Kum.width.mean", "Osa_Bar.width.mean", "Osa_Nor.width.mean" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Osa_Oki.width.mean", "Osa_Kum.width.mean", "Osa_Bar.width.mean", "Osa_Nor.width.mean")) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(span = 0.25, se = FALSE) +
  ggtitle("Osaka: Chromosomal position vs. aligned segment width")

p3 <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Bar_Oki.width.mean", "Bar_Kum.width.mean", "Bar_Osa.width.mean", "Bar_Aom.width.mean" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Bar_Oki.width.mean", "Bar_Kum.width.mean", "Bar_Osa.width.mean", "Bar_Aom.width.mean")) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(span = 0.25, se = FALSE) +
  ggtitle("Barcelona: Chromosomal position vs. aligned segment width")

ggpubr::ggarrange(p1, p2, p3, ncol=1)
```


## Breakpoints

```{r chromosome_plots_bp, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# get_bps returns a GRanges, but we need mcols to use grWindows.
# Add some mcols to the output of get_bps.
get_bps_mcol <- function(gr) {
  gr <- get_bps(gr)
  gr$breaks <- "break"
  gr
}

```

```{r chromosome_plots_bp_oki, dependson="chromosome_plots_bp", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Osa), windowSize=windowSize, meta_prefix = 'bp.Oki_Osa.')))

mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Bar), windowSize=windowSize, meta_prefix = 'bp.Oki_Bar.')))

mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Kum), windowSize=windowSize, meta_prefix = 'bp.Oki_Kum.')))

mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Aom), windowSize=windowSize, meta_prefix = 'bp.Oki_Aom.')))

mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Nor), windowSize=windowSize, meta_prefix = 'bp.Oki_Nor.')))
```

```{r chromosome_plots_bp_oki_1, dependson="chromosome_plots_bp", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1_ysr <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) |>
  tidyr::pivot_longer(c("bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. number of breakpoints")

p1_ysr

p1_noysr <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. number of breakpoints")
```

```{r chromosome_plots_bp_osa, dependson="chromosome_plots_bp", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Oki), windowSize=windowSize, meta_prefix = 'bp.Osa_Oki.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Bar), windowSize=windowSize, meta_prefix = 'bp.Osa_Bar.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Kum), windowSize=windowSize, meta_prefix = 'bp.Osa_Kum.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Aom), windowSize=windowSize, meta_prefix = 'bp.Osa_Aom.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Nor), windowSize=windowSize, meta_prefix = 'bp.Osa_Nor.')))
```

```{r chromosome_plots_bp_osa_1, dependson="chromosome_plots_bp_osa", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p2_ysr <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) |>
  tidyr::pivot_longer(c("bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Osaka: Chromosomal position vs. number of breakpoints")

p2_ysr

p2_noysr <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(span = 0.25, se = FALSE) +
    ggtitle("Osaka: Chromosomal position vs. number of breakpoints")

```


```{r chromosome_plots_bp_bar, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Oki), windowSize=windowSize, meta_prefix = 'bp.Bar_Oki.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Osa), windowSize=windowSize, meta_prefix = 'bp.Bar_Osa.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Kum), windowSize=windowSize, meta_prefix = 'bp.Bar_Kum.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Aom), windowSize=windowSize, meta_prefix = 'bp.Bar_Aom.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Nor), windowSize=windowSize, meta_prefix = 'bp.Bar_Nor.')))
```

```{r chromosome_plots_bp_bar_1, dependson="chromosome_plots_bp_bar", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p3_ysr <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) |>
  tidyr::pivot_longer(c("bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count" )) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(span = 0.25, se = FALSE) +
  ggtitle("Barcelona: Chromosomal position vs. number of breakpoints")

p3_ysr 

p3_noysr <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count" )) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(span = 0.25, se = FALSE) +
  ggtitle("Barcelona: Chromosomal position vs. number of breakpoints")

```

```{r chromosome_plots_bp_joined, dependson=c("chromosome_plots_bp_oki", "chromosome_plots_bp_osa", "chromosome_plots_bp_bar"), fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
ggpubr::ggarrange(p1_ysr, p2_ysr, p3_ysr, ncol=1)
ggpubr::ggarrange(p1_noysr, p2_noysr, p3_noysr, ncol=1)
```
# Session information

```{r sessionInfo}
sessionInfo()
```
