---
title: "Chromosome-scale feature density plots"
author: 
 - "Michael Mansfield"
 - "Charles Plessy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Chromosome-scale feature density plots}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(cache = TRUE)
```

# Introduction

In this vignette, we develop some of the materials needed to produce plots that summarize genomic data at chromosome-scale.

The core functions used here are maintained in our _GenomicBreaks_ R package,
which is fully documented at: <https://oist.github.io/GenomicBreaks>.

## Load packages

# Load R packages and data

```{r load_packages_and_data}
library('OikScrambling') |> suppressPackageStartupMessages()
library('viridis') |> suppressPackageStartupMessages()
library('dplyr') |> suppressPackageStartupMessages()
requireNamespace('patchwork') |> suppressPackageStartupMessages()
library("patchwork")

genomes <- OikScrambling:::loadAllGenomes()
transcripts <- OikScrambling:::loadAllTranscriptsGR()
reps <- OikScrambling:::loadAllRepeats()
load("BreakPoints.Rdata")
```

### Creating chromosome-scale plots
In this vignette, we create the plumbing needed to generate summary plots that show windows of genomic features on a chromosome scale. 

#### Summarizing genomic windows of annotated `GRanges` objects
Below, I create some of the functions needed to produce these plots.

```{r chromosome_plot_helper_fxns}
# Make a BSgenome into a GRanges object, one range per chromosome
BSgenomeToGR <- function(genome) {
  as(seqinfo(genome), "GRanges")
}
(tmp <- genomes$Oki |> BSgenomeToGR())

# Make a genome into a bunch of windows of length N
genomeGRToWindows <- function(gr, windowSize=10000) {
  tile(gr, w=windowSize) |> unlist() |> unname()
}
(tmp <- genomes$Oki |> BSgenomeToGR() |> genomeGRToWindows())

# Match ranges to a single bin by reducing each range to its center.
# Returns the index of the matched bins
matchToOneBin <- function(ranges, bins) {
  # Ensure that a range overlaps with only one bin
  safeRanges <- ranges |> resize(1, fix="center")
  o <- findOverlaps(safeRanges, bins, type = 'within')
  subjectHits(o)
}
(matchToOneBin(transcripts$Oki, tmp)) |> head()

# Apply one function to data grouped by bins, and array it according to the
# original bins (that is, insert NAs for bins that contained no data).
# This function only works as intended if the bins fully tile the genome.
binApply <- function(f, bins, data, index, ...) {
  res <- rep(NA, length(bins))
  vals <-tapply(data, index, f, ...)
  res[as.numeric(names(vals))] <- unname(vals)
  res
}
(binApply(mean, tmp, transcripts$Oki$dNdS_GUIDANCE2, matchToOneBin(transcripts$Oki, tmp), na.rm = TRUE)) |> head(100)

# Intersect each genome window with some other GRange (meta, as in metadata)
metaPerWindow <- function(genomeWindows, meta, meta_prefix=NULL) {
  idx <- matchToOneBin(meta, genomeWindows)
  for (name in colnames(mcols(meta))) {
    data <- mcols(meta)[[name]]
    if(is.numeric(data)) {
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".count")]]       <- binApply(length, genomeWindows, data, idx)
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".median")]]      <- binApply(median, genomeWindows, data, idx, na.rm = TRUE)
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".mean")]]        <- binApply(mean,   genomeWindows, data, idx, na.rm = TRUE)
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".sd")]]          <- binApply(sd,     genomeWindows, data, idx, na.rm = TRUE)
    } else if(is.character(data) ) {
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".count")]]       <- binApply(length, genomeWindows, data, idx)
    } else if(is.factor(data)) {
      mcols(genomeWindows)[[paste0(meta_prefix, name, ".count.total")]] <- binApply(length, genomeWindows, data, idx)
      if(length(levels(data)) > 1) {
        binned_list_of_tables <- binApply(table,  genomeWindows, data, idx)
        binned_DataFrame <- do.call(rbind, binned_list_of_tables) |> DataFrame()
        colnames(binned_DataFrame) <- paste0(meta_prefix, name, ".", colnames(binned_DataFrame), ".count.total")
        mcols(genomeWindows) <- cbind(mcols(genomeWindows), binned_DataFrame)
      }
    }
  }
  genomeWindows
}
(metaPerWindow(genomeWindows=tmp, transcripts$Oki, meta_prefix = 'transcripts.'))

# Function to do everything
grWindows <- function(genome, meta, windowSize=10000, meta_prefix=NULL){
  genome <- BSgenomeToGR(genome)                                 # Make GRanges from genome
  genome <- genomeGRToWindows(genome, windowSize=windowSize)     # Make a windowed GRange from chromosome GRanges
  mpw <-    metaPerWindow(genome, meta, meta_prefix=meta_prefix) # Make metadata column windows and intersect
  mpw
}
```

## Chromosome plots
### Transcripts
In the following sections we annotate and initialize the `chrw` object, which is a `SimpleList()` containing windowed chromosome regions. 

We use these windows to evaluate the distribution of various features across the chromosomes. This section concerns one-dimensional features.

```{r chromosome_plots_transcripts}
windowSize <- 2e5
bins_Oki <- genomes$Oki |> BSgenomeToGR() |> genomeGRToWindows(windowSize)
bins_Osa <- genomes$Osa |> BSgenomeToGR() |> genomeGRToWindows(windowSize)
bins_Bar <- genomes$Bar |> BSgenomeToGR() |> genomeGRToWindows(windowSize)

transcripts$Oki$tx_len <- width(transcripts$Oki)
transcripts$Osa$tx_len <- width(transcripts$Osa)
transcripts$Bar$tx_len <- width(transcripts$Bar)

chrw <- SimpleList()
chrw$Oki <- grWindows(genome=genomes$Oki, meta=transcripts$Oki, windowSize=windowSize, meta_prefix = 'transcripts.')
chrw$Oki <- flagLongShort(chrw$Oki, longShort$OKI2018.I69)
chrw$Osa <- grWindows(genome=genomes$Osa, meta=transcripts$Osa, windowSize=windowSize, meta_prefix = 'transcripts.')
chrw$Osa <- flagLongShort(chrw$Osa, longShort$OSKA2016v1.9)
chrw$Bar <- grWindows(genome=genomes$Bar, meta=transcripts$Bar, windowSize=windowSize, meta_prefix = 'transcripts.')
chrw$Bar <- flagLongShort(chrw$Bar, longShort$Bar2.p4)
```

```{r chromosome_plots_transcripts_oki, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Create a list of plots. We save some of these plots for faceting later.
plots <- list()
plots$Oki <- list()
plots$Oki$transcripts <- list()

plots$Oki$transcripts$density_1 <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.tx_name.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula = y ~ x) + ggtitle(paste("Okinawa: transcripts per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$OKI2018.I69 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Oki$transcripts$density_2 <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_id.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: transcripts per window (size = ", windowSize/1000, " kbp)", sep=""))
plots$Oki$transcripts$dnds_1 <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula = y ~ x) + ggtitle(paste("Okinawa: median dN/dS per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$OKI2018.I69 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Oki$transcripts$dnds_2 <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: median dN/dS per window (size = ", windowSize/1000, " kbp)", sep=""))

plots$Oki$transcripts$lengths <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_len.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: median transcript length per window (size = ", windowSize/1000, " kbp)", sep=""))

lapply(plots$Oki$transcripts, \(x) x)
```


```{r chromosome_plots_transcripts_osa, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Osa <- list()
plots$Osa$transcripts <- list()

plots$Osa$transcripts$density_1 <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.tx_name.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula = y ~ x) + ggtitle(paste("Osaka: transcripts per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$OSKA2016v1.9 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Osa$transcripts$density_2 <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_id.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Osaka: transcripts per window (size = ", windowSize/1000, " kbp)", sep=""))

plots$Osa$transcripts$dnds_1 <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula = y ~ x) + ggtitle(paste("Osaka: median dN/dS per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$OSKA2016v1.9 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Osa$transcripts$dnds_2 <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Osaka: median dN/dS per window (size = ", windowSize/1000, " kbp)", sep=""))

plots$Osa$transcripts$lengths <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_len.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Osaka: median transcript length per window (size = ", windowSize/1000, " kbp)", sep=""))

lapply(plots$Osa$transcripts, \(x) x)
```

```{r chromosome_plots_transcripts_bar, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Bar <- list()
plots$Bar$transcripts <- list()

plots$Bar$transcripts$density_1 <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.tx_name.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula = y ~ x) + ggtitle(paste("Barcelona: transcripts per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$Bar2.p4 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Bar$transcripts$density_2 <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_id.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Barcelona: transcripts per window (size = ", windowSize/1000, " kbp)", sep=""))

plots$Bar$transcripts$dnds_1 <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula = y ~ x) + ggtitle(paste("Barcelona: median dN/dS per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$Bar2.p4 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Bar$transcripts$dnds_2 <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.dNdS_PRANK.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Barcelona: median dN/dS per window (size = ", windowSize/1000, " kbp)", sep=""))

plots$Bar$transcripts$lengths <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=transcripts.tx_len.median, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess',  method.args = list(family = "symmetric")) + ggtitle(paste("Barcelona: median transcript length per window (size = ", windowSize/1000, " kbp)", sep=""))

lapply(plots$Bar$transcripts, \(x) x)
```

#### Transcripts: joined gene density plot
```{r chromosome_plots_transcripts_joined, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
layout <- "
A
B
C
"

( plots$Oki$transcripts$density_2 / plots$Osa$transcripts$density_2 / plots$Bar$transcripts$density_2 ) +   plot_layout(design = layout, guides = 'collect')
```

#### Transcripts: joined dN/dS plot
```{r chromosome_plots_transcripts_dnds_joined, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
layout <- "
A
B
C
"

( plots$Oki$transcripts$dnds_2 / plots$Osa$transcripts$dnds_2 / plots$Bar$transcripts$dnds_2 ) +   plot_layout(design = layout, guides = 'collect')
```


### Operons
In this section we add operon data to the `chrw` object. Since the windows are the same for every genome, this is simple accomplished with a `cbind` operation to the windowed object's `mcols`.

```{r chromosome_plots_operons, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Since the coordinates of the windows are the same, you can simply cbind more metadata columns to the data frame.
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=OikScrambling:::calcOperons(transcripts$Oki, window = 500), windowSize=windowSize, meta_prefix = 'operons.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=OikScrambling:::calcOperons(transcripts$Osa, window = 500), windowSize=windowSize, meta_prefix = 'operons.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=OikScrambling:::calcOperons(transcripts$Bar, window = 500), windowSize=windowSize, meta_prefix = 'operons.')))
```


```{r chromosome_plots_operons_oki, dependson="chromosome_plots_operons", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Oki$operons <- list()

plots$Oki$operons$density_1 <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=operons.n.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula=y~x) + ggtitle(paste("Okinawa: operons per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$OKI2018.I69 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Oki$operons$density_2 <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=operons.n.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: operons per window (size = ", windowSize/1000, " kbp)", sep=""))

plots$Oki$operons$length_1 <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=operons.n.mean) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula=y~x) + ggtitle(paste("Okinawa: mean operon size per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$OKI2018.I69 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Oki$operons$length_2 <- ggplot(chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=operons.n.mean, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric"), se=FALSE) + ggtitle(paste("Okinawa: mean operon size per window (size = ", windowSize/1000, " kbp)", sep=""))

lapply(plots$Oki$operons, \(x) x)
```


```{r chromosome_plots_operons_osa, dependson="chromosome_plots_transcripts", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Osa$operons <- list()

plots$Osa$operons$density_1 <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=operons.n.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula = y ~ x) + ggtitle(paste("Osaka: operons per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$OSKA2016v1.9 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Osa$operons$density_2 <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=operons.n.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric")) + ggtitle(paste("Osaka: operons per window (size = ", windowSize/1000, " kbp)", sep=""))

plots$Osa$operons$length_1 <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=operons.n.mean) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula=y~x) + ggtitle(paste("Osaka: mean operon size per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$OSKA2016v1.9 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Osa$operons$length_2 <- ggplot(chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=operons.n.mean, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric"), se=FALSE) + ggtitle(paste("Osaka: mean operon size per window (size = ", windowSize/1000, " kbp)", sep=""))

lapply(plots$Osa$operons, \(x) x)
```

```{r chromosome_plots_operons_bar, dependson="chromosome_plots_operons", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Bar$operons <- list()

plots$Bar$operons$density_1 <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=operons.n.count) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula=y~x) + ggtitle(paste("Barcelona: operons per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$Bar2.p4 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Bar$operons$density_2 <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=operons.n.count, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric")) + ggtitle(paste("Barcelona: operons per window (size = ", windowSize/1000, " kbp)", sep=""))

plots$Bar$operons$length_1 <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=operons.n.mean) + geom_point() + facet_wrap(~seqnames) + geom_smooth(method='loess', formula=y~x) + ggtitle(paste("Barcelona: mean operon size per window (size = ", windowSize/1000, " kbp)", sep="")) + geom_vline( data = (longShort$Bar2.p4 -1) |> GenomicBreaks::cleanGaps() |> as.data.frame(), aes(xintercept = start), col = "gray")

plots$Bar$operons$length_2 <- ggplot(chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=operons.n.mean, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric"), se=FALSE) + ggtitle(paste("Barcelona: mean operon size per window (size = ", windowSize/1000, " kbp)", sep=""))

lapply(plots$Bar$operons, \(x) x)
```

#### Operons: joined operon density plot
```{r chromosome_plots_operons_joined, dependson="chromosome_plots_operons", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}

layout <- "
A
B
C
"

(plots$Oki$operons$density_2 / plots$Osa$operons$density_2 / plots$Bar$operons$density_2 ) + plot_layout(design = layout, guides = 'collect')
(plots$Oki$operons$length_2 / plots$Osa$operons$length_2 / plots$Bar$operons$length_2 ) + plot_layout(design = layout, guides = 'collect')

(plots$Oki$transcripts$lengths / plots$Oki$operons$length_2 / plots$Osa$transcripts$lengths / plots$Osa$operons$length_2 / plots$Bar$transcripts$lengths   / plots$Bar$operons$length_2 ) + plot_layout(ncol=2, nrow=3)

```

### Repeats
Like operons, it is simple to add repeat density, since the X coordinates (`start` and `end` positions of a `GRanges` are shared).

```{r chromosome_plots_repeats, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=reps$Oki, windowSize=windowSize, meta_prefix = 'repeats.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=reps$Osa, windowSize=windowSize, meta_prefix = 'repeats.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=reps$Bar, windowSize=windowSize, meta_prefix = 'repeats.')))

```

We can use the `[mcol].counts.tables` metadata columns to calculate statistics between different genomic regions.

```{r chromosome_plots_repeats_oki, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Oki$repeats <- list()

plots$Oki$repeats$density_1 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=repeats.Class.count.total) + geom_point() + facet_wrap(seqnames~Arm) + geom_smooth(method='loess', formula = y ~ x) + scale_y_log10()

plots$Oki$repeats$density_2 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) ) + aes(x=start, y=repeats.type.count.total, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric")) + ggtitle(paste("Okinawa: repeats per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('repeats')

plots$Oki$repeats$classes <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.ltr.1.count.total", "repeats.Class.rnd.count.total", "repeats.Class.SINE.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.ltr.1.count.total", "repeats.Class.rnd.count.total", "repeats.Class.SINE.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Density of repeat classes across genome")
lapply(plots$Oki$repeats, \(x) x)
```

```{r chromosome_plots_repeats_osa, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Osa$repeats <- list()

plots$Osa$repeats$density_1 <- ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=repeats.Class.count.total) + geom_point() + facet_wrap(seqnames~Arm) + geom_smooth(method='loess', formula = y ~ x) + scale_y_log10()

plots$Osa$repeats$density_2 <- ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=repeats.type.count.total, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric")) + ggtitle(paste("Osa: repeats per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('repeats')

plots$Osa$repeats$classes <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.rnd.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.rnd.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Osaka: distribution of repeat classes across genome")

lapply(plots$Osa$repeats, \(x) x)
```



```{r chromosome_plots_repeats_bar, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Bar$repeats <- list()

plots$Bar$repeats$density_1 <- ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + aes(x=start, y=repeats.Class.count.total) + geom_point() + facet_wrap(seqnames~Arm) + geom_smooth(method='loess', formula = y ~ x) + scale_y_log10()

plots$Bar$repeats$density_2 <- ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) ) + aes(x=start, y=repeats.type.count.total, col = Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(method='loess', formula = y ~ x, method.args = list(family = "symmetric")) + ggtitle(paste("Barcelona: repeats per window (size = ", windowSize/1000, " kbp)", sep="")) + ylab('repeats')

plots$Bar$repeats$classes <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.rnd.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("repeats.Class.count.total", "repeats.Class.LowComplexity.count.total", "repeats.Class.rnd.count.total", "repeats.Class.tandem.count.total", "repeats.Class.unknown.count.total" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Barcelona: distribution of repeat classes across genome")

lapply(plots$Bar$repeats, \(x) x)
```

```{r chromosome_plots_repeats_joined, dependson="chromosome_plots_repeats", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
layout <- "
A
B
C
"

(plots$Oki$repeats$density_2 / plots$Osa$repeats$density_2 / plots$Bar$repeats$density_2) +   plot_layout(design = layout, guides = 'collect')
((plots$Oki$repeats$density_2 + scale_y_log10()) + (plots$Osa$repeats$density_2 + scale_y_log10()) + (plots$Bar$repeats$density_2 + scale_y_log10() )) + plot_layout(design = layout, guides = 'collect')

```

## `GenomicBreaks` objects: Scores and widths

In the next sections, we explore the distribution of alignment parameters (scores and widths) for different populations vs. chromosomal position.

```{r chromosome_plots_scores, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Okinawa
chrw$Oki$Oki_Osa.score.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Osa), matchToOneBin(gbs$Oki_Osa, bins_Oki))
chrw$Oki$Oki_Aom.score.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Aom), matchToOneBin(gbs$Oki_Aom, bins_Oki))
chrw$Oki$Oki_Bar.score.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Bar), matchToOneBin(gbs$Oki_Bar, bins_Oki))
chrw$Oki$Oki_Nor.score.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Nor), matchToOneBin(gbs$Oki_Nor, bins_Oki))

# Osaka
chrw$Osa$Osa_Oki.score.mean <- binApply(mean, na.rm = T, bins_Osa, score(gbs$Osa_Oki), matchToOneBin(gbs$Osa_Oki, bins_Osa))
chrw$Osa$Osa_Kum.score.mean <- binApply(mean, na.rm = T, bins_Osa, score(gbs$Osa_Kum), matchToOneBin(gbs$Osa_Kum, bins_Osa))
chrw$Osa$Osa_Bar.score.mean <- binApply(mean, na.rm = T, bins_Osa, score(gbs$Osa_Bar), matchToOneBin(gbs$Osa_Bar, bins_Osa))
chrw$Osa$Osa_Nor.score.mean <- binApply(mean, na.rm = T, bins_Osa, score(gbs$Osa_Nor), matchToOneBin(gbs$Osa_Nor, bins_Osa))

# Barcelona
chrw$Bar$Bar_Osa.score.mean <- binApply(mean, na.rm = T, bins_Bar, score(gbs$Bar_Osa), matchToOneBin(gbs$Bar_Osa, bins_Bar))
chrw$Bar$Bar_Aom.score.mean <- binApply(mean, na.rm = T, bins_Bar, score(gbs$Bar_Aom), matchToOneBin(gbs$Bar_Aom, bins_Bar))
chrw$Bar$Bar_Oki.score.mean <- binApply(mean, na.rm = T, bins_Bar, score(gbs$Bar_Oki), matchToOneBin(gbs$Bar_Oki, bins_Bar))
chrw$Bar$Bar_Kum.score.mean <- binApply(mean, na.rm = T, bins_Bar, score(gbs$Bar_Kum), matchToOneBin(gbs$Bar_Kum, bins_Bar))
```

```{r chromosome_plots_scores_oki, dependson="chromosome_plots_scores", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Oki$scores <- list()

plots$Oki$scores$oki_osa <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Osa.score.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth(method='loess', formula = y ~ x) +
  scale_y_log10() +
  ggtitle("Oki-Osa: Chromosomal position vs. alignment score")

plots$Oki$scores$oki_bar <-  ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Bar.score.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth(method='loess', formula = y ~ x) +
  scale_y_log10() + 
  ggtitle("Oki-Bar: Chromosomal position vs. alignment score")

plots$Oki$scores$combined <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki_Osa.score.mean", "Oki_Aom.score.mean", "Oki_Bar.score.mean", "Oki_Nor.score.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Oki_Osa.score.mean", "Oki_Aom.score.mean", "Oki_Bar.score.mean", "Oki_Nor.score.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. alignment score")

plots$Osa$scores <- list()
plots$Osa$scores$combined <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Osa_Oki.score.mean", "Osa_Kum.score.mean", "Osa_Bar.score.mean", "Osa_Nor.score.mean" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Osa_Oki.score.mean", "Osa_Kum.score.mean", "Osa_Bar.score.mean", "Osa_Nor.score.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Osaka: Chromosomal position vs. alignment score")

plots$Bar$scores <- list()
plots$Bar$scores$combined <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Bar_Oki.score.mean", "Bar_Kum.score.mean", "Bar_Osa.score.mean", "Bar_Aom.score.mean" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Bar_Oki.score.mean", "Bar_Kum.score.mean", "Bar_Osa.score.mean", "Bar_Aom.score.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Barcelona: Chromosomal position vs. alignment score")

lapply(plots$Oki$scores, \(x) x)
plots$Osa$scores$combined
plots$Bar$scores$combined
```


```{r chromosome_plots_scores_joined, dependson="chromosome_plots_scores", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
layout <- "
A
B
C
"

(plots$Oki$scores$combined / plots$Osa$scores$combined / plots$Bar$scores$combined) +   plot_layout(design = layout, guides = 'collect')
```


In the following section I divide the scores by the widths of the alignments, creating a scaled score.

```{r chromosome_plots_scores_norm, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
chrw$Oki$Oki_Osa.score.norm.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Osa)/lengths(gbs$Oki_Osa), matchToOneBin(gbs$Oki_Osa, bins_Oki))
chrw$Oki$Oki_Aom.score.norm.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Aom)/lengths(gbs$Oki_Aom), matchToOneBin(gbs$Oki_Aom, bins_Oki))
chrw$Oki$Oki_Bar.score.norm.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Bar)/lengths(gbs$Oki_Bar), matchToOneBin(gbs$Oki_Bar, bins_Oki))
chrw$Oki$Oki_Nor.score.norm.mean <- binApply(mean, na.rm = T, bins_Oki, score(gbs$Oki_Nor)/lengths(gbs$Oki_Nor), matchToOneBin(gbs$Oki_Nor, bins_Oki))

ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Osa.score.norm.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth(method='loess', formula = y ~ x) +
  scale_y_log10() +
  ggtitle("Oki-Osa: Chromosomal position vs. scaled alignment score")

ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Bar.score.norm.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth(method='loess', formula = y ~ x) +
  scale_y_log10() +
  ggtitle("Oki-Bar: Chromosomal position vs. scaled alignment score")


chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki_Osa.score.norm.mean", "Oki_Aom.score.norm.mean", "Oki_Bar.score.norm.mean", "Oki_Nor.score.norm.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Oki_Osa.score.norm.mean", "Oki_Aom.score.norm.mean", "Oki_Bar.score.norm.mean", "Oki_Nor.score.norm.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. scaled alignment score")

```

## Alignment widths

```{r chromosome_plots_widths, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Okinawa
chrw$Oki$Oki_Osa.width.mean <- binApply(mean, na.rm = T, bins_Oki, width(gbs$Oki_Osa), matchToOneBin(gbs$Oki_Osa, bins_Oki))
chrw$Oki$Oki_Aom.width.mean <- binApply(mean, na.rm = T, bins_Oki, width(gbs$Oki_Aom), matchToOneBin(gbs$Oki_Aom, bins_Oki))
chrw$Oki$Oki_Bar.width.mean <- binApply(mean, na.rm = T, bins_Oki, width(gbs$Oki_Bar), matchToOneBin(gbs$Oki_Bar, bins_Oki))
chrw$Oki$Oki_Nor.width.mean <- binApply(mean, na.rm = T, bins_Oki, width(gbs$Oki_Nor), matchToOneBin(gbs$Oki_Nor, bins_Oki))

# Osaka
chrw$Osa$Osa_Oki.width.mean <- binApply(mean, na.rm = T, bins_Osa, width(gbs$Osa_Oki), matchToOneBin(gbs$Osa_Oki, bins_Osa))
chrw$Osa$Osa_Kum.width.mean <- binApply(mean, na.rm = T, bins_Osa, width(gbs$Osa_Kum), matchToOneBin(gbs$Osa_Kum, bins_Osa))
chrw$Osa$Osa_Bar.width.mean <- binApply(mean, na.rm = T, bins_Osa, width(gbs$Osa_Bar), matchToOneBin(gbs$Osa_Bar, bins_Osa))
chrw$Osa$Osa_Nor.width.mean <- binApply(mean, na.rm = T, bins_Osa, width(gbs$Osa_Nor), matchToOneBin(gbs$Osa_Nor, bins_Osa))

# Barcelona
chrw$Bar$Bar_Osa.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Osa), matchToOneBin(gbs$Bar_Osa, bins_Bar))
chrw$Bar$Bar_Aom.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Aom), matchToOneBin(gbs$Bar_Aom, bins_Bar))
chrw$Bar$Bar_Oki.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Oki), matchToOneBin(gbs$Bar_Oki, bins_Bar))
chrw$Bar$Bar_Kum.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Kum), matchToOneBin(gbs$Bar_Kum, bins_Bar))
```

```{r chromosome_plots_widths_oki, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
plots$Oki$widths <- list()
plots$Oki$widths$widths_1 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Osa.width.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth(method='loess', formula = y ~ x) +
  scale_y_log10() +
  ggtitle("Oki-Osa: Chromosomal position vs. aligned segment width")

plots$Oki$widths$widths_2 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) +
  aes(x=start, y=Oki_Osa.width.mean) +
  geom_point() +
  facet_wrap(~seqnames) +
  geom_smooth(method='loess', formula = y ~ x) +
  scale_y_log10() +
  ggtitle("Oki-Bar: Chromosomal position vs. aligned segment width")

plots$Oki$widths$combined <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki_Osa.width.mean", "Oki_Aom.width.mean", "Oki_Bar.width.mean", "Oki_Nor.width.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Oki_Osa.width.mean", "Oki_Aom.width.mean", "Oki_Bar.width.mean", "Oki_Nor.width.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. aligned segment width")

plots$Osa$widths <- list()
plots$Osa$widths$combined <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Osa_Oki.width.mean", "Osa_Kum.width.mean", "Osa_Bar.width.mean", "Osa_Nor.width.mean" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Osa_Oki.width.mean", "Osa_Kum.width.mean", "Osa_Bar.width.mean", "Osa_Nor.width.mean")) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
  ggtitle("Osaka: Chromosomal position vs. aligned segment width")

plots$Bar$widths <- list()
plots$Bar$widths$combined <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Bar_Oki.width.mean", "Bar_Kum.width.mean", "Bar_Osa.width.mean", "Bar_Aom.width.mean" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("Bar_Oki.width.mean", "Bar_Kum.width.mean", "Bar_Osa.width.mean", "Bar_Aom.width.mean")) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
  ggtitle("Barcelona: Chromosomal position vs. aligned segment width")

lapply(plots$Oki$widths, \(x) x)
plots$Osa$widths$combined
plots$Bar$widths$combined
```

```{r chromosome_plots_widths_joined, dependson="chromosome_plots_widths", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
layout <- "
A
B
C
"

(plots$Oki$widths$combined / plots$Osa$widths$combined / plots$Bar$widths$combined) +   plot_layout(design = layout, guides = 'collect')
```

## Breakpoints

```{r chromosome_plots_bp, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# get_bps returns a GRanges, but we need mcols to use grWindows.
# Add some mcols to the output of get_bps.
get_bps_mcol <- function(gr) {
  gr <- get_bps(gr)
  gr$breaks <- "break"
  gr
}

```

```{r chromosome_plots_bp_oki, dependson="chromosome_plots_bp", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Osa), windowSize=windowSize, meta_prefix = 'bp.Oki_Osa.')))

mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Bar), windowSize=windowSize, meta_prefix = 'bp.Oki_Bar.')))

mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Kum), windowSize=windowSize, meta_prefix = 'bp.Oki_Kum.')))

mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Aom), windowSize=windowSize, meta_prefix = 'bp.Oki_Aom.')))

mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Nor), windowSize=windowSize, meta_prefix = 'bp.Oki_Nor.')))

# And a mean count
chrw$Oki$bp.mean.breaks <- chrw$Oki |> as.data.frame()|> select("bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count") |> as.data.frame() |> rowMeans()
```

```{r chromosome_plots_bp_oki_1, dependson="chromosome_plots_bp", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1_ysr <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) |>
  tidyr::pivot_longer(c("bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. number of breakpoints")

p1_ysr

p1_noysr <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Okinawa: Chromosomal position vs. number of breakpoints")
```

```{r chromosome_plots_bp_osa, dependson="chromosome_plots_bp", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Oki), windowSize=windowSize, meta_prefix = 'bp.Osa_Oki.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Bar), windowSize=windowSize, meta_prefix = 'bp.Osa_Bar.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Kum), windowSize=windowSize, meta_prefix = 'bp.Osa_Kum.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Aom), windowSize=windowSize, meta_prefix = 'bp.Osa_Aom.')))

mcols(chrw$Osa) <- cbind(mcols(chrw$Osa), mcols(grWindows(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Nor), windowSize=windowSize, meta_prefix = 'bp.Osa_Nor.')))

chrw$Osa$bp.mean.breaks <- chrw$Osa |> as.data.frame() |> select("bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count") |> as.data.frame() |> rowMeans()
```

```{r chromosome_plots_bp_osa_1, dependson="chromosome_plots_bp_osa", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p2_ysr <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) |>
  tidyr::pivot_longer(c("bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Osaka: Chromosomal position vs. number of breakpoints")

p2_ysr

p2_noysr <- chrw$Osa |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count" )) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Osaka: Chromosomal position vs. number of breakpoints")

```


```{r chromosome_plots_bp_bar, dependson="chromosome_plot_helper_fxns", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Oki), windowSize=windowSize, meta_prefix = 'bp.Bar_Oki.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Osa), windowSize=windowSize, meta_prefix = 'bp.Bar_Osa.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Kum), windowSize=windowSize, meta_prefix = 'bp.Bar_Kum.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Aom), windowSize=windowSize, meta_prefix = 'bp.Bar_Aom.')))

mcols(chrw$Bar) <- cbind(mcols(chrw$Bar), mcols(grWindows(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Nor), windowSize=windowSize, meta_prefix = 'bp.Bar_Nor.')))
```

```{r chromosome_plots_bp_bar_1, dependson="chromosome_plots_bp_bar", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p3_ysr <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) |>
  tidyr::pivot_longer(c("bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count" )) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
  ggtitle("Barcelona: Chromosomal position vs. number of breakpoints")

p3_ysr 

p3_noysr <- chrw$Bar |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count" )) |>
  plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count" )) |>
  ggplot() +
  aes(start, value, col = name) +
  facet_wrap(~seqnames, nrow = 1) +
  geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
  ggtitle("Barcelona: Chromosomal position vs. number of breakpoints")

```

```{r chromosome_plots_bp_joined, dependson=c("chromosome_plots_bp_oki", "chromosome_plots_bp_osa", "chromosome_plots_bp_bar"), fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
layout <- "
A
B
C
"

(p1_ysr / p2_ysr / p3_ysr ) +   plot_layout(design = layout, guides = 'collect')
(p1_noysr / p2_noysr / p3_noysr ) +   plot_layout(design = layout, guides = 'collect')
```

#### Plotting everything on one plot
Note that there are two Y axes for these plots. Repeat and gene density are shown on the same Y axis (left), while dN/dS is shown on the second Y axis (right).


```{r chromosome_plots_bp_oki_2, dependson=c("chromosome_plots_bp_oki", "chromosome_plots_bp_osa", "chromosome_plots_bp_bar"), fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tmp <- chrw$Oki |> as.data.frame() |> plyranges::filter(seqnames %in% c("chr1", "chr2", "PAR", "XSR"))

tmp$bp.mean.breaks <- tmp |> select("bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count") |> as.data.frame() |> rowMeans()

scaleFactor <- max(tmp$bp.mean.breaks,na.rm=T)/max(tmp$transcripts.dNdS_PRANK.median, na.rm=T)
p1 <- ggplot(tmp, aes(x=start)) +
  geom_smooth(method='loess', formula = y ~ x, aes(y=bp.mean.breaks, color="genes")) +
  geom_smooth(method='loess', formula = y ~ x, aes(y=transcripts.dNdS_PRANK.median*scaleFactor, color="dN/dS")) +
  geom_smooth(method='loess', formula = y ~ x, aes(y=repeats.Class.count.total, color="repeats")) +
  scale_y_continuous(name="breakpoints/repeats per window", sec.axis=sec_axis(~./scaleFactor, name="median dN/dS per window")) +
  ggtitle("Okinawa: Breakpoints, genes, and dN/dS per window") + 
  facet_wrap(~seqnames, nrow=1)

p1

# And a correlation plot
ggplot(tmp) + aes(x=bp.mean.breaks, y=transcripts.dNdS_GUIDANCE2.median) + geom_point() + geom_smooth(method='lm', col='red') + ggpubr::stat_cor() + facet_wrap(seqnames~Arm, ncol=2, scales = 'free') + ggtitle('Okinawa: Breakpoints vs. dN/dS') + ylab('dN/dS')
```


```{r chromosome_plots_bp_osa_2, dependson=c("chromosome_plots_bp_oki", "chromosome_plots_bp_osa", "chromosome_plots_bp_bar"), fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tmp <- chrw$Osa |> as.data.frame() |> plyranges::filter(seqnames %in% c("Chr1", "Chr2", "PAR", "XSR"))

tmp$bp.mean.breaks <- tmp |> select("bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count",  "bp.Osa_Nor.breaks.count") |> as.data.frame() |> rowMeans()

scaleFactor <- max(tmp$bp.mean.breaks,na.rm=T)/max(tmp$transcripts.dNdS_PRANK.median, na.rm=T)

p2 <- ggplot(tmp, aes(x=start)) +
  geom_smooth(method='loess', formula = y ~ x, aes(y=bp.mean.breaks, color="genes")) +
  geom_smooth(method='loess', formula = y ~ x, aes(y=transcripts.dNdS_PRANK.median*scaleFactor, color="dN/dS")) +
  geom_smooth(method='loess', formula = y ~ x, aes(y=repeats.Class.count.total, color="repeats")) +
  scale_y_continuous(name="breakpoints/repeats per window", sec.axis=sec_axis(~./scaleFactor, name="median dN/dS per window")) +
  ggtitle("Osaka: Breakpoints, genes, and dN/dS per window") + 
  facet_wrap(~seqnames, nrow=1)

p2

ggplot(tmp) + aes(x=bp.mean.breaks, y=transcripts.dNdS_GUIDANCE2.median) + geom_point() + geom_smooth(method='lm', col='red') + ggpubr::stat_cor() + facet_wrap(seqnames~Arm, ncol=2, scales = 'free') + ggtitle('Osaka: Breakpoints vs. dN/dS') + ylab('dN/dS')
```

```{r chromosome_plots_bp_bar_2, dependson=c("chromosome_plots_bp_oki", "chromosome_plots_bp_osa", "chromosome_plots_bp_bar"), fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tmp <- chrw$Bar |> as.data.frame() |> plyranges::filter(seqnames %in% c("Chr1", "Chr2", "PAR", "XSR"))

tmp$bp.mean.breaks <- tmp |> select("bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count",  "bp.Bar_Nor.breaks.count") |> as.data.frame() |> rowMeans()

scaleFactor <- max(tmp$bp.mean.breaks,na.rm=T)/max(tmp$transcripts.dNdS_PRANK.median, na.rm=T)

p3 <- ggplot(tmp, aes(x=start)) +
  geom_smooth(method='loess', formula = y ~ x,aes(y=bp.mean.breaks, color="genes")) +
  geom_smooth(method='loess', formula = y ~ x,aes(y=transcripts.dNdS_PRANK.median*scaleFactor, color="dN/dS")) +
  geom_smooth(method='loess', formula = y ~ x,aes(y=repeats.Class.count.total, color="repeats")) +
  scale_y_continuous(name="breakpoints/repeats per window", sec.axis=sec_axis(~./scaleFactor, name="median dN/dS per window")) +
  ggtitle("Barcelona: Breakpoints, genes, and dN/dS per window") + 
  facet_wrap(~seqnames, nrow=1)

p3

ggplot(tmp) + aes(x=bp.mean.breaks, y=transcripts.dNdS_GUIDANCE2.median) + geom_point() + geom_smooth(method='lm', col='red') + ggpubr::stat_cor() + facet_wrap(seqnames~Arm, ncol=2, scales = 'free') + ggtitle('Barcelona: Breakpoints vs. dN/dS') + ylab('dN/dS')
```

```{r chromosome_plots_bp_joined_2, dependson=c("chromosome_plots_bp_oki", "chromosome_plots_bp_osa", "chromosome_plots_bp_bar"), fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
(p1 / p2 / p3 ) + plot_layout(design = layout, guides = 'collect')
```

## Alignment categories
The 4 alignment categories we describe can vary between genomic compartments. In this section, I analyse their distributions.

```{r chromosome_plots_categories_helper_fxns}
makeWholeGenomeObject <- function(aligned, unaligned) {
  isolated  <- aligned[aligned$nonCoa]
  collinear <- aligned[!aligned$nonCoa]
  breakpoint  <- unaligned[unaligned$nonCoa]
  bridge    <- unaligned[!unaligned$nonCoa]

  # Let's keep only the common metadata
  mcols(isolated)  <- mcols(isolated)[ ,colnames(mcols(unaligned))]
  mcols(collinear) <- mcols(collinear)[,colnames(mcols(unaligned))]
  
  # Let's replace the `nonCoa` flag by a `aligned`, `unaligned`, `syntenic` flag.
  isolated$nonCoa <- collinear$nonCoa <- breakpoint$nonCoa <- bridge$nonCoa <- NULL
  isolated$type   <- 'isolated'
  collinear$type  <- 'collinear'
  breakpoint$type <- 'breakpoint'
  bridge$type     <- 'bridge'

  wholeGenome <- c(isolated, collinear, breakpoint, bridge) |> sort(ignore.strand = TRUE)
  wholeGenome$type <- factor(wholeGenome$type)
  wholeGenome
}


wgo2 <- mapply(makeWholeGenomeObject, gbs, unal) |> SimpleList()
# We really only care about "type" for this analysis, so let's strip out all the other metadata.
wgo2 <- lapply(wgo2, function(g) { 
  gr <- g
  mcols(gr) <- mcols(gr)[["type"]]
  colnames(mcols(gr)) <- "type"
  gr
}) |> SimpleList()
```

```{r chromosome_plots_categories_oki, dependson=c("chromosome_plots_bp_oki", "chromosome_plots_bp_osa", "chromosome_plots_bp_bar"), fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
# Okinawa
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=wgo2$Oki_Osa, windowSize=windowSize, meta_prefix = 'ac.Oki_Osa.')))
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=wgo2$Oki_Bar, windowSize=windowSize, meta_prefix = 'ac.Oki_Bar.')))
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=wgo2$Oki_Kum, windowSize=windowSize, meta_prefix = 'ac.Oki_Kum.')))
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=wgo2$Oki_Aom, windowSize=windowSize, meta_prefix = 'ac.Oki_Aom.')))
mcols(chrw$Oki) <- cbind(mcols(chrw$Oki), mcols(grWindows(genome=genomes$Oki, meta=wgo2$Oki_Nor, windowSize=windowSize, meta_prefix = 'ac.Oki_Nor.')))

chrw$Oki$ac.isolated.mean <- chrw$Oki |> as.data.frame()|> select("ac.Oki_Osa.type.isolated.count.total", "ac.Oki_Bar.type.isolated.count.total", "ac.Oki_Kum.type.isolated.count.total", "ac.Oki_Aom.type.isolated.count.total", "ac.Oki_Nor.type.isolated.count.total") |> as.data.frame() |> rowMeans()
chrw$Oki$ac.collinear.mean <- chrw$Oki |> as.data.frame()|> select("ac.Oki_Osa.type.collinear.count.total", "ac.Oki_Bar.type.collinear.count.total", "ac.Oki_Kum.type.collinear.count.total", "ac.Oki_Aom.type.collinear.count.total", "ac.Oki_Nor.type.collinear.count.total") |> as.data.frame() |> rowMeans()
chrw$Oki$ac.breakpoint.mean <- chrw$Oki |> as.data.frame()|> select("ac.Oki_Osa.type.breakpoint.count.total", "ac.Oki_Bar.type.breakpoint.count.total", "ac.Oki_Kum.type.breakpoint.count.total", "ac.Oki_Aom.type.breakpoint.count.total", "ac.Oki_Nor.type.breakpoint.count.total") |> as.data.frame() |> rowMeans()
chrw$Oki$ac.bridge.mean <- chrw$Oki |> as.data.frame()|> select("ac.Oki_Osa.type.bridge.count.total", "ac.Oki_Bar.type.bridge.count.total", "ac.Oki_Kum.type.bridge.count.total", "ac.Oki_Aom.type.bridge.count.total", "ac.Oki_Nor.type.bridge.count.total") |> as.data.frame() |> rowMeans()

plots$Oki$categories <- list()

plots$Oki$categories$oki_osa_classes <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "ac.Oki_Osa.type.isolated.count.total", "ac.Oki_Osa.type.collinear.count.total", "ac.Oki_Osa.type.breakpoint.count.total", "ac.Oki_Osa.type.bridge.count.total" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
  tidyr::pivot_longer(c("ac.Oki_Osa.type.isolated.count.total", "ac.Oki_Osa.type.collinear.count.total", "ac.Oki_Osa.type.breakpoint.count.total", "ac.Oki_Osa.type.bridge.count.total")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Okinawa-Osaka: Distribution of alignment categories across genome")

plots$Oki$categories$oki_osa_overall <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Arm", "ac.Oki_Osa.type.isolated.count.total", "ac.Oki_Osa.type.collinear.count.total", "ac.Oki_Osa.type.breakpoint.count.total", "ac.Oki_Osa.type.bridge.count.total" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
 tidyr::pivot_longer(c("ac.Oki_Osa.type.isolated.count.total", "ac.Oki_Osa.type.collinear.count.total", "ac.Oki_Osa.type.breakpoint.count.total", "ac.Oki_Osa.type.bridge.count.total")) |>
  ggplot() +
    aes(value, paste0(seqnames, "_", Arm),  fill = name) +
    geom_bar(position='fill', stat='identity') +
    ylab("chr and arm") +
    ggtitle("Okinawa-Osaka: Overall proportion of alignment categories across genome")

plots$Oki$categories$oki_osa_by_chr_position <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Arm", "ac.Oki_Osa.type.isolated.count.total", "ac.Oki_Osa.type.collinear.count.total", "ac.Oki_Osa.type.breakpoint.count.total", "ac.Oki_Osa.type.bridge.count.total" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
 tidyr::pivot_longer(c("ac.Oki_Osa.type.isolated.count.total", "ac.Oki_Osa.type.collinear.count.total", "ac.Oki_Osa.type.breakpoint.count.total", "ac.Oki_Osa.type.bridge.count.total")) |>
  ggplot() +
    aes(as.factor(start), value, fill = name) +
    facet_wrap(~seqnames, nrow = 1, scales='free') +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
    geom_bar(position='fill', stat='identity') +
    xlab("start") +
    ggtitle("Okinawa-Osaka: Proportion of alignment categories by position")

# Next, plot the means of ALL populations vs. Oki.
plots$Oki$categories$mean_classes <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Arm", "ac.isolated.mean", "ac.collinear.mean", "ac.breakpoint.mean", "ac.bridge.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
 tidyr::pivot_longer(c("ac.isolated.mean", "ac.collinear.mean", "ac.breakpoint.mean", "ac.bridge.mean")) |>
  ggplot() +
    aes(start, value, col = name) +
    facet_wrap(~seqnames, nrow = 1) +
    geom_smooth(method='loess', formula = y ~ x, span = 0.25, se = FALSE) +
    ggtitle("Okinawa vs. all: Distribution of alignment categories across genome")

plots$Oki$categories$mean_classes_overall <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Arm", "ac.isolated.mean", "ac.collinear.mean", "ac.breakpoint.mean", "ac.bridge.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
 tidyr::pivot_longer(c("ac.isolated.mean", "ac.collinear.mean", "ac.breakpoint.mean", "ac.bridge.mean")) |>
  ggplot() +
    aes(value, paste0(seqnames, "_", Arm),  fill = name) +
    geom_bar(position='fill', stat='identity') +
    ylab("chr and arm") +
    ggtitle("Okinawa vs. all: Overall proportion of alignment categories across genome")

plots$Oki$categories$mean_classes_by_chr_position <- chrw$Oki |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Arm", "ac.isolated.mean", "ac.collinear.mean", "ac.breakpoint.mean", "ac.bridge.mean" )) |>
  plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |>
 tidyr::pivot_longer(c("ac.isolated.mean", "ac.collinear.mean", "ac.breakpoint.mean", "ac.bridge.mean")) |>
  ggplot() +
    aes(as.factor(start), value, fill = name) +
    facet_wrap(~seqnames, nrow = 1, scales='free') +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
    geom_bar(position='fill', stat='identity') +
    xlab("start") +
    ggtitle("Okinawa vs. all: Proportion of alignment categories by position")

lapply(plots$Oki$categories, \(x) x)
```


## Statistics
```{r chromosome_plots_statistics_1, dependson="chromosome_plots_operons", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
append_width <- function(gr) {
  gr$w <- width(gr)
  gr
}

ggplot(transcripts$Oki |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) |> flagLongShort(longShort$OKI2018.I69) |> append_width() |> as.data.frame() ) +
  aes(x=paste0(seqnames, "_", Arm), y=w, fill=seqnames) +
  scale_y_continuous(trans='log10') +
  ylab('transcript width') +
  xlab(NULL) +
  theme(axis.text.x=element_text(angle=90)) +
  ggtitle('Okinawa: transcript width vs. chr+arm') + 
  geom_boxplot()

p1 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + geom_boxplot(aes(x=paste0(seqnames, "_", Arm), y=transcripts.tx_len.median, fill=seqnames)) + ylab("median transcript length") + xlab(NULL) + theme(axis.text.x=element_text(angle=90)) + ggtitle("Okinawa: transcript length by chr and arm")

p2 <- ggplot(as.data.frame(chrw$Oki) |> plyranges::filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR')) ) + geom_boxplot(aes(x=paste0(seqnames, "_", Arm), y=operons.n.mean, fill=seqnames)) + ylab("operon size") + xlab(NULL) + theme(axis.text.x=element_text(angle=90)) + ggtitle("Okinawa: operon size by chr and arm")

p3 <- ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + geom_boxplot(aes(x=paste0(seqnames, "_", Arm), y=transcripts.tx_len.median, fill=seqnames)) + ylab("median transcript length") + xlab(NULL) + theme(axis.text.x=element_text(angle=90)) + ggtitle("Osa: transcript length by chr and arm")

p4 <- ggplot(as.data.frame(chrw$Osa) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + geom_boxplot(aes(x=paste0(seqnames, "_", Arm), y=operons.n.mean, fill=seqnames)) + ylab("operon size") + xlab(NULL) + theme(axis.text.x=element_text(angle=90)) + ggtitle("Osa: operon size by chr and arm")

p5 <- ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + geom_boxplot(aes(x=paste0(seqnames, "_", Arm), y=transcripts.tx_len.median, fill=seqnames)) + ylab("median transcript length") + xlab(NULL) + theme(axis.text.x=element_text(angle=90)) + ggtitle("Bar: transcript length by chr and arm")

p6 <- ggplot(as.data.frame(chrw$Bar) |> plyranges::filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR')) ) + geom_boxplot(aes(x=paste0(seqnames, "_", Arm), y=operons.n.mean, fill=seqnames)) + ylab("operon size") + xlab(NULL) + theme(axis.text.x=element_text(angle=90)) + ggtitle("Bar: operon size by chr and arm")

layout <- "
AA
BB
CC
"

(p1 + p2 + p3 + p4 + p5 + p6 ) +   plot_layout(ncol=2, nrow=3)
```


```{r chromosome_plots_statistics_2_oki, dependson="chromosome_plots_operons", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
library(ggpubr)
tb <- chrw$Oki |> as.data.frame() |> filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR', 'YSR'))
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Arm)
plots$Oki$boxplots <- list()
plots$Oki$boxplots$operon_density <- ggboxplot(tb, x='seqnames_arm', y='operons.n.count', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('operon density per window') +
  stat_compare_means(comparison=list(c('chr1_short', 'chr1_long'), c('chr2_short', 'chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Okinawa: operon density by chr+arm')

plots$Oki$boxplots$operon_size <- ggboxplot(tb, x='seqnames_arm', y='operons.n.mean', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('operon size per window') +
  stat_compare_means(comparison=list(c('chr1_short', 'chr1_long'), c('chr2_short', 'chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Okinawa: operon size by chr+arm')

plots$Oki$boxplots$transcript_density <- ggboxplot(tb, x='seqnames_arm', y='transcripts.tx_id.count', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('transcript size per window') +
  stat_compare_means(comparison=list(c('chr1_short', 'chr1_long'), c('chr2_short', 'chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Okinawa: transcript density by chr+arm')

plots$Oki$boxplots$transcript_size <- ggboxplot(tb, x='seqnames_arm', y='transcripts.tx_len.median', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('transcript size per window') +
  stat_compare_means(comparison=list(c('chr1_short', 'chr1_long'), c('chr2_short', 'chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Okinawa: transcript size by chr+arm')

plots$Oki$boxplots$dnds <- ggboxplot(tb, x='seqnames_arm', y='transcripts.dNdS_GUIDANCE2.median', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('median dN/dS per window') +
  stat_compare_means(comparison=list(c('chr1_short', 'chr1_long'), c('chr2_short', 'chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Okinawa: median dN/dS by chr+arm')

lapply(plots$Oki$boxplots, \(x) x)
```

```{r chromosome_plots_statistics_2_osa, dependson="chromosome_plots_operons", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tb <- chrw$Osa |> as.data.frame() |> filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR'))
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Arm)
plots$Osa$boxplots <- list()
plots$Osa$boxplots$operon_density <- ggboxplot(tb, x='seqnames_arm', y='operons.n.count', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('operon density per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Osaka: operon density by chr+arm')

plots$Osa$boxplots$operon_size <- ggboxplot(tb, x='seqnames_arm', y='operons.n.mean', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('operon size per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Osaka: operon size by chr+arm')

plots$Osa$boxplots$transcript_density <- ggboxplot(tb, x='seqnames_arm', y='transcripts.tx_id.count', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('transcript size per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Osaka: transcript density by chr+arm')

plots$Osa$boxplots$transcript_size <- ggboxplot(tb, x='seqnames_arm', y='transcripts.tx_len.median', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('transcript size per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Osaka: transcript size by chr+arm')

plots$Osa$boxplots$dnds <- ggboxplot(tb, x='seqnames_arm', y='transcripts.dNdS_GUIDANCE2.median', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('median dN/dS per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Osaka: median dN/dS by chr+arm')

lapply(plots$Osa$boxplots, \(x) x)
```


```{r chromosome_plots_statistics_2_bar, dependson="chromosome_plots_operons", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tb <- chrw$Bar |> as.data.frame() |> filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR', 'YSR'))
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Arm)
plots$Bar$boxplots <- list()
plots$Bar$boxplots$operon_density <- ggboxplot(tb, x='seqnames_arm', y='operons.n.count', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('operon density per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Barcelona: operon density by chr+arm')

plots$Bar$boxplots$operon_size <- ggboxplot(tb, x='seqnames_arm', y='operons.n.mean', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('operon size per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Barcelona: operon size by chr+arm')

plots$Bar$boxplots$transcript_density <- ggboxplot(tb, x='seqnames_arm', y='transcripts.tx_id.count', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('transcript size per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Barcelona: transcript density by chr+arm')

plots$Bar$boxplots$transcript_size <- ggboxplot(tb, x='seqnames_arm', y='transcripts.tx_len.median', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('transcript size per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Barcelona: transcript size by chr+arm')

plots$Bar$boxplots$dnds <- ggboxplot(tb, x='seqnames_arm', y='transcripts.dNdS_GUIDANCE2.median', color='seqnames', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('median dN/dS per window') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'YSR_YSR'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long'))) +
  ggtitle('Barcelona: median dN/dS by chr+arm')

lapply(plots$Bar$boxplots, \(x) x)
detach('package:ggpubr')
```

## Transforming to a single coordinate system
In this section, I create some utility functions to scale genomic ranges to the same coordinate system, then calculate averages of some statistics over the windows in these regions.

First, create some helper functions for tile-based indexing.

```{r chromosome_plots_tiles_helper_fxns}
genomeGRToTiles <- function(gr, n=100) {
  tile(gr, n=n) |> unlist() |> unname()
}

grTiles <- function(genome, meta, n=100, meta_prefix=NULL){
  genome <- BSgenomeToGR(genome)                                 # Make GRanges from genome
  genome <- genomeGRToTiles(genome, n=n)                         # Make a tiled GRange from chromosome GRanges
  mpw <-    metaPerWindow(genome, meta, meta_prefix=meta_prefix) # Make metadata column windows and intersect
  mpw
}
```


```{r chromosome_plots_tiles_calcs, dependson="chromosome_plot_tiles_helper_fxns"}
nTiles <- 100
tiles_Oki <- genomes$Oki |> BSgenomeToGR() |> genomeGRToTiles(nTiles)
tiles_Osa <- genomes$Osa |> BSgenomeToGR() |> genomeGRToTiles(nTiles)
tiles_Bar <- genomes$Bar |> BSgenomeToGR() |> genomeGRToTiles(nTiles)

chrt <- SimpleList()
chrt$Oki <- grTiles(genome=genomes$Oki, meta=transcripts$Oki, n=nTiles, meta_prefix = 'transcripts.')
chrt$Oki <- flagLongShort(chrt$Oki, longShort$OKI2018.I69)
chrt$Osa <- grTiles(genome=genomes$Osa, meta=transcripts$Osa, n=nTiles, meta_prefix = 'transcripts.')
chrt$Osa <- flagLongShort(chrt$Osa, longShort$OSKA2016v1.9)
chrt$Bar <- grTiles(genome=genomes$Bar, meta=transcripts$Bar, n=nTiles, meta_prefix = 'transcripts.')
chrt$Bar <- flagLongShort(chrt$Bar, longShort$Bar2.p4)

# Operons
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=OikScrambling:::calcOperons(transcripts$Oki, window = 500), n=nTiles, meta_prefix = 'operons.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=OikScrambling:::calcOperons(transcripts$Osa, window = 500), n=nTiles, meta_prefix = 'operons.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=OikScrambling:::calcOperons(transcripts$Bar, window = 500), n=nTiles, meta_prefix = 'operons.')))

# Repeats
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=reps$Oki, n=nTiles, meta_prefix = 'repeats.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=reps$Osa, n=nTiles, meta_prefix = 'repeats.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=reps$Bar, n=nTiles, meta_prefix = 'repeats.')))

# Widths
chrt$Oki$Oki_Osa.width.mean <- binApply(mean, na.rm = T, tiles_Oki, width(gbs$Oki_Osa), matchToOneBin(gbs$Oki_Osa, tiles_Oki))
chrt$Oki$Oki_Aom.width.mean <- binApply(mean, na.rm = T, tiles_Oki, width(gbs$Oki_Aom), matchToOneBin(gbs$Oki_Aom, tiles_Oki))
chrt$Oki$Oki_Bar.width.mean <- binApply(mean, na.rm = T, tiles_Oki, width(gbs$Oki_Bar), matchToOneBin(gbs$Oki_Bar, tiles_Oki))
chrt$Oki$Oki_Nor.width.mean <- binApply(mean, na.rm = T, tiles_Oki, width(gbs$Oki_Nor), matchToOneBin(gbs$Oki_Nor, tiles_Oki))
chrt$Osa$Osa_Oki.width.mean <- binApply(mean, na.rm = T, tiles_Osa, width(gbs$Osa_Oki), matchToOneBin(gbs$Osa_Oki, tiles_Osa))
chrt$Osa$Osa_Kum.width.mean <- binApply(mean, na.rm = T, tiles_Osa, width(gbs$Osa_Kum), matchToOneBin(gbs$Osa_Kum, tiles_Osa))
chrt$Osa$Osa_Bar.width.mean <- binApply(mean, na.rm = T, tiles_Osa, width(gbs$Osa_Bar), matchToOneBin(gbs$Osa_Bar, tiles_Osa))
chrt$Osa$Osa_Nor.width.mean <- binApply(mean, na.rm = T, tiles_Osa, width(gbs$Osa_Nor), matchToOneBin(gbs$Osa_Nor, tiles_Osa))
chrt$Bar$Bar_Osa.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Osa), matchToOneBin(gbs$Bar_Osa, bins_Bar))
chrt$Bar$Bar_Aom.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Aom), matchToOneBin(gbs$Bar_Aom, bins_Bar))
chrt$Bar$Bar_Oki.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Oki), matchToOneBin(gbs$Bar_Oki, bins_Bar))
chrt$Bar$Bar_Kum.width.mean <- binApply(mean, na.rm = T, bins_Bar, width(gbs$Bar_Kum), matchToOneBin(gbs$Bar_Kum, bins_Bar))

# Breakpoints
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Osa), n=nTiles, meta_prefix = 'bp.Oki_Osa.')))
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Bar), n=nTiles, meta_prefix = 'bp.Oki_Bar.')))
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Kum), n=nTiles, meta_prefix = 'bp.Oki_Kum.')))
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Aom), n=nTiles, meta_prefix = 'bp.Oki_Aom.')))
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=get_bps_mcol(gbs$Oki_Nor), n=nTiles, meta_prefix = 'bp.Oki_Nor.')))
chrt$Oki$bp.mean.breaks <- chrt$Oki |> as.data.frame()|> select("bp.Oki_Osa.breaks.count", "bp.Oki_Bar.breaks.count", "bp.Oki_Kum.breaks.count", "bp.Oki_Aom.breaks.count",  "bp.Oki_Nor.breaks.count") |> as.data.frame() |> rowMeans()

mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Oki), n=nTiles, meta_prefix = 'bp.Osa_Oki.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Bar), n=nTiles, meta_prefix = 'bp.Osa_Bar.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Kum), n=nTiles, meta_prefix = 'bp.Osa_Kum.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Aom), n=nTiles, meta_prefix = 'bp.Osa_Aom.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=get_bps_mcol(gbs$Osa_Nor), n=nTiles, meta_prefix = 'bp.Osa_Nor.')))
chrt$Osa$bp.mean.breaks <- chrt$Osa |> as.data.frame()|> select("bp.Osa_Oki.breaks.count", "bp.Osa_Bar.breaks.count", "bp.Osa_Kum.breaks.count", "bp.Osa_Aom.breaks.count", "bp.Osa_Nor.breaks.count") |> as.data.frame() |> rowMeans()

mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Oki), n=nTiles, meta_prefix = 'bp.Bar_Oki.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Osa), n=nTiles, meta_prefix = 'bp.Bar_Osa.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Kum), n=nTiles, meta_prefix = 'bp.Bar_Kum.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Aom), n=nTiles, meta_prefix = 'bp.Bar_Aom.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=get_bps_mcol(gbs$Bar_Nor), n=nTiles, meta_prefix = 'bp.Bar_Nor.')))
chrt$Bar$bp.mean.breaks <- chrt$Bar |> as.data.frame()|> select("bp.Bar_Oki.breaks.count", "bp.Bar_Osa.breaks.count", "bp.Bar_Kum.breaks.count", "bp.Bar_Aom.breaks.count", "bp.Bar_Nor.breaks.count") |> as.data.frame() |> rowMeans()

# Alignment categories
# Okinawa
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=wgo2$Oki_Osa, n=nTiles, meta_prefix = 'ac.Oki_Osa.')))
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=wgo2$Oki_Bar, n=nTiles, meta_prefix = 'ac.Oki_Bar.')))
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=wgo2$Oki_Kum, n=nTiles, meta_prefix = 'ac.Oki_Kum.')))
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=wgo2$Oki_Aom, n=nTiles, meta_prefix = 'ac.Oki_Aom.')))
mcols(chrt$Oki) <- cbind(mcols(chrt$Oki), mcols(grTiles(genome=genomes$Oki, meta=wgo2$Oki_Nor, n=nTiles, meta_prefix = 'ac.Oki_Nor.')))
chrt$Oki$ac.isolated.mean <- chrt$Oki |> as.data.frame()|> select("ac.Oki_Osa.type.isolated.count.total", "ac.Oki_Bar.type.isolated.count.total", "ac.Oki_Kum.type.isolated.count.total", "ac.Oki_Aom.type.isolated.count.total", "ac.Oki_Nor.type.isolated.count.total") |> as.data.frame() |> rowMeans()
chrt$Oki$ac.collinear.mean <- chrt$Oki |> as.data.frame()|> select("ac.Oki_Osa.type.collinear.count.total", "ac.Oki_Bar.type.collinear.count.total", "ac.Oki_Kum.type.collinear.count.total", "ac.Oki_Aom.type.collinear.count.total", "ac.Oki_Nor.type.collinear.count.total") |> as.data.frame() |> rowMeans()
chrt$Oki$ac.breakpoint.mean <- chrt$Oki |> as.data.frame()|> select("ac.Oki_Osa.type.breakpoint.count.total", "ac.Oki_Bar.type.breakpoint.count.total", "ac.Oki_Kum.type.breakpoint.count.total", "ac.Oki_Aom.type.breakpoint.count.total", "ac.Oki_Nor.type.breakpoint.count.total") |> as.data.frame() |> rowMeans()
chrt$Oki$ac.bridge.mean <- chrt$Oki |> as.data.frame()|> select("ac.Oki_Osa.type.bridge.count.total", "ac.Oki_Bar.type.bridge.count.total", "ac.Oki_Kum.type.bridge.count.total", "ac.Oki_Aom.type.bridge.count.total", "ac.Oki_Nor.type.bridge.count.total") |> as.data.frame() |> rowMeans()

mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=wgo2$Osa_Oki, n=nTiles, meta_prefix = 'ac.Osa_Oki.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=wgo2$Osa_Bar, n=nTiles, meta_prefix = 'ac.Osa_Bar.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=wgo2$Osa_Kum, n=nTiles, meta_prefix = 'ac.Osa_Kum.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=wgo2$Osa_Aom, n=nTiles, meta_prefix = 'ac.Osa_Aom.')))
mcols(chrt$Osa) <- cbind(mcols(chrt$Osa), mcols(grTiles(genome=genomes$Osa, meta=wgo2$Osa_Nor, n=nTiles, meta_prefix = 'ac.Osa_Nor.')))
chrt$Osa$ac.isolated.mean <- chrt$Osa |> as.data.frame()|> select("ac.Osa_Oki.type.isolated.count.total", "ac.Osa_Bar.type.isolated.count.total", "ac.Osa_Kum.type.isolated.count.total", "ac.Osa_Aom.type.isolated.count.total", "ac.Osa_Nor.type.isolated.count.total") |> as.data.frame() |> rowMeans()
chrt$Osa$ac.collinear.mean <- chrt$Osa |> as.data.frame()|> select("ac.Osa_Oki.type.collinear.count.total", "ac.Osa_Bar.type.collinear.count.total", "ac.Osa_Kum.type.collinear.count.total", "ac.Osa_Aom.type.collinear.count.total", "ac.Osa_Nor.type.collinear.count.total") |> as.data.frame() |> rowMeans()
chrt$Osa$ac.breakpoint.mean <- chrt$Osa |> as.data.frame()|> select("ac.Osa_Oki.type.breakpoint.count.total", "ac.Osa_Bar.type.breakpoint.count.total", "ac.Osa_Kum.type.breakpoint.count.total", "ac.Osa_Aom.type.breakpoint.count.total", "ac.Osa_Nor.type.breakpoint.count.total") |> as.data.frame() |> rowMeans()
chrt$Osa$ac.bridge.mean <- chrt$Osa |> as.data.frame()|> select("ac.Osa_Oki.type.bridge.count.total", "ac.Osa_Bar.type.bridge.count.total", "ac.Osa_Kum.type.bridge.count.total", "ac.Osa_Aom.type.bridge.count.total", "ac.Osa_Nor.type.bridge.count.total") |> as.data.frame() |> rowMeans()

mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=wgo2$Bar_Oki, n=nTiles, meta_prefix = 'ac.Bar_Oki.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=wgo2$Bar_Osa, n=nTiles, meta_prefix = 'ac.Bar_Osa.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=wgo2$Bar_Kum, n=nTiles, meta_prefix = 'ac.Bar_Kum.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=wgo2$Bar_Aom, n=nTiles, meta_prefix = 'ac.Bar_Aom.')))
mcols(chrt$Bar) <- cbind(mcols(chrt$Bar), mcols(grTiles(genome=genomes$Bar, meta=wgo2$Bar_Nor, n=nTiles, meta_prefix = 'ac.Bar_Nor.')))
chrt$Bar$ac.isolated.mean <- chrt$Bar |> as.data.frame()|> select("ac.Bar_Oki.type.isolated.count.total", "ac.Bar_Osa.type.isolated.count.total", "ac.Bar_Kum.type.isolated.count.total", "ac.Bar_Aom.type.isolated.count.total", "ac.Bar_Nor.type.isolated.count.total") |> as.data.frame() |> rowMeans()
chrt$Bar$ac.collinear.mean <- chrt$Bar |> as.data.frame()|> select("ac.Bar_Oki.type.collinear.count.total", "ac.Bar_Osa.type.collinear.count.total", "ac.Bar_Kum.type.collinear.count.total", "ac.Bar_Aom.type.collinear.count.total", "ac.Bar_Nor.type.collinear.count.total") |> as.data.frame() |> rowMeans()
chrt$Bar$ac.breakpoint.mean <- chrt$Bar |> as.data.frame()|> select("ac.Bar_Oki.type.breakpoint.count.total", "ac.Bar_Osa.type.breakpoint.count.total", "ac.Bar_Kum.type.breakpoint.count.total", "ac.Bar_Aom.type.breakpoint.count.total", "ac.Bar_Nor.type.breakpoint.count.total") |> as.data.frame() |> rowMeans()
chrt$Bar$ac.bridge.mean <- chrt$Bar |> as.data.frame()|> select("ac.Bar_Oki.type.bridge.count.total", "ac.Bar_Osa.type.bridge.count.total", "ac.Bar_Kum.type.bridge.count.total", "ac.Bar_Aom.type.bridge.count.total", "ac.Bar_Nor.type.bridge.count.total") |> as.data.frame() |> rowMeans()






# Now, combining it all into a single unified tiled chromosome object.
chrt_unified <- data.frame(seqnames=c(rep('Chr1', nTiles), rep('Chr2', nTiles), rep('PAR', nTiles), rep('XSR', nTiles)), start=c(seq(1, nTiles), seq(1, nTiles), seq(1, nTiles), seq(1, nTiles)), end=c(seq(1, nTiles)+1, seq(1, nTiles)+1, seq(1, nTiles)+1, seq(1, nTiles)+1) )  |> GRanges()
seqlevels(chrt_unified) <- c('Chr1', 'Chr2', 'PAR', 'XSR')

# Add a prefix to mcols to prevent collisions between colnames between similar objects.
prefix_cbind_mcols <- function(gr1, gr2, prefix) {
  gr <- gr1
  q_df <- mcols(gr1)
  t_df <- mcols(gr2)
  colnames(t_df) <- lapply(colnames(t_df), function(x) paste0(prefix, '.', x)) |> unlist()
  mcols(gr) <- cbind(q_df, t_df)
  gr
}
chrt_unified <- prefix_cbind_mcols(chrt_unified, chrt$Oki |> as.data.frame() |> filter(seqnames %in% c('chr1', 'chr2', 'PAR', 'XSR')) |> GRanges() , prefix='Oki')
chrt_unified <- prefix_cbind_mcols(chrt_unified, chrt$Osa |> as.data.frame() |> filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |> GRanges() , prefix='Osa')
chrt_unified <- prefix_cbind_mcols(chrt_unified, chrt$Bar |> as.data.frame() |> filter(seqnames %in% c('Chr1', 'Chr2', 'PAR', 'XSR')) |> GRanges() , prefix='Bar')

# Making some mean values across all species
chrt_unified$transcripts.count.mean <- data.frame(chrt_unified$Oki.transcripts.tx_id.count, chrt_unified$Osa.transcripts.tx_id.count, chrt_unified$Bar.transcripts.tx_id.count) |> rowMeans()

chrt_unified$transcripts.length.mean <- data.frame(chrt_unified$Oki.transcripts.tx_len.mean, chrt_unified$Osa.transcripts.tx_len.mean, chrt_unified$Bar.transcripts.tx_len.mean) |> rowMeans()

chrt_unified$operons.count.mean <- data.frame(chrt_unified$Oki.operons.n.count, chrt_unified$Osa.operons.n.count, chrt_unified$Bar.operons.n.count) |> rowMeans()
chrt_unified$operons.length.mean <- data.frame(chrt_unified$Oki.operons.n.mean, chrt_unified$Osa.operons.n.mean, chrt_unified$Bar.operons.n.mean) |> rowMeans()

chrt_unified$transcripts.dNdS.mean <- data.frame(chrt_unified$Oki.transcripts.dNdS_GUIDANCE2.median, chrt_unified$Osa.transcripts.dNdS_GUIDANCE2.median, chrt_unified$Bar.transcripts.dNdS_GUIDANCE2.median) |> rowMeans()

chrt_unified$repeats.count.mean <- data.frame(chrt_unified$Oki.repeats.type.count.total, chrt_unified$Osa.repeats.type.count.total, chrt_unified$Bar.repeats.type.count.total) |> rowMeans()

chrt_unified$breakpoints.count.mean <- data.frame(chrt_unified$Oki.bp.mean.breaks, chrt_unified$Osa.bp.mean.breaks, chrt_unified$Bar.bp.mean.breaks) |> rowMeans()

chrt_unified$categories.isolated.mean <- data.frame(chrt_unified$Oki.ac.isolated.mean, chrt_unified$Osa.ac.isolated.mean, chrt_unified$Bar.ac.isolated.mean) |> rowMeans()
chrt_unified$categories.collinear.mean <- data.frame(chrt_unified$Oki.ac.collinear.mean, chrt_unified$Osa.ac.collinear.mean, chrt_unified$Bar.ac.collinear.mean) |> rowMeans()
chrt_unified$categories.breakpoint.mean <- data.frame(chrt_unified$Oki.ac.breakpoint.mean, chrt_unified$Osa.ac.breakpoint.mean, chrt_unified$Bar.ac.breakpoint.mean) |> rowMeans()
chrt_unified$categories.bridge.mean <- data.frame(chrt_unified$Oki.ac.bridge.mean, chrt_unified$Osa.ac.bridge.mean, chrt_unified$Bar.ac.bridge.mean) |> rowMeans()

```


```{r chromosome_plots_tiles_1, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
ggplot(chrt_unified |> as.data.frame()) +
  geom_smooth(aes(x=start, y=transcripts.count.mean/max(transcripts.count.mean, na.rm=T), col='transcripts'), se=FALSE) +
  geom_smooth(aes(x=start, y=repeats.count.mean/max(repeats.count.mean, na.rm = T), col='repeats'), se=FALSE) +
  geom_smooth(aes(x=start, y=transcripts.dNdS.mean/max(transcripts.dNdS.mean, na.rm = T), col='dN/dS'), se=FALSE) +
  geom_smooth(aes(x=start, y=operons.count.mean/max(operons.count.mean, na.rm = T), col='operons'), se=FALSE) +
  geom_smooth(aes(x=start, y=breakpoints.count.mean/max(breakpoints.count.mean, na.rm = T), col='breakpoints'), se=FALSE) +
  ylab('pct') +
  facet_wrap(~seqnames, nrow=1) +
  ggtitle('Tiles (averages across Oki, Osa, Bar): Properties across genome')
```


What features correlate with breakpoints?

```{r chromosome_plots_tiles_2, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p1 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.dNdS.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('dN/dS')

p2 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.count.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('transcripts')

p3 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=operons.count.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('transcripts')

p4 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=repeats.count.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('repeats')

( p1 / p2 / p3 / p4)
```

What alignment categories correlate with alignment categories?

```{r chromosome_plots_tiles_3, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p1 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=categories.isolated.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('isolated')

p2 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=categories.collinear.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('collinear')

p3 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=categories.breakpoint.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('breakpoint')

p4 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=categories.bridge.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('bridge')

( p1 / p2 / p3 / p4)
```



```{r chromosome_plots_tiles_5, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
ggplot(chrt_unified |> as.data.frame()) +
  geom_smooth(aes(x=start, y=categories.isolated.mean,  col='isolated'), se=FALSE) +
  geom_smooth(aes(x=start, y=categories.collinear.mean, col='collinear'), se=FALSE) +
  geom_smooth(aes(x=start, y=categories.breakpoint.mean,  col='breakpoint'), se=FALSE) +
  geom_smooth(aes(x=start, y=categories.bridge.mean,    col='bridge'), se=FALSE) +
  ylab('pct') +
  facet_wrap(~seqnames, nrow=1) +
  ggtitle('Tiles (averages across Oki, Osa, Bar): Alignment categories by location')

```


Gene size and density vs. operon size and density.

```{r chromosome_plots_tiles_5_2, dependson="chromosome_plots_tiles_calcs", fig.height=10, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p0 <- ggplot(tb) +
  geom_smooth(aes(x=start, y=transcripts.count.mean/max(transcripts.count.mean, na.rm = T),  col='gene density'), formula=y~x, method='loess', se=FALSE) +
  geom_smooth(aes(x=start, y=transcripts.length.mean/max(transcripts.length.mean, na.rm = T),  col='gene length'), formula=y~x, method='loess', se=FALSE) +
  geom_smooth(aes(x=start, y=operons.length.mean/max(operons.length.mean,na.rm=T),  col='operon length'), formula=y~x, method='loess', se=FALSE) +
  geom_smooth(aes(x=start, y=operons.count.mean/max(operons.count.mean,na.rm=T),  col='operon density'), formula=y~x, method='loess', se=FALSE) +
  geom_smooth(aes(x=start, y=transcripts.dNdS.mean/max(transcripts.dNdS.mean, na.rm = T),  col='dN/dS'), formula=y~x, method='loess', se=FALSE) +
  geom_smooth(aes(x=start, y=breakpoints.count.mean/max(breakpoints.count.mean, na.rm = T),  col='breakpoints'), formula=y~x, method='loess', se=FALSE) +
  ylab("Normalized mean per tile across all species") +
  facet_wrap(~seqnames, nrow=1)
p0

p1 <- ggplot(tb) +
  geom_smooth(aes(x=start, y=transcripts.count.mean/max(transcripts.count.mean, na.rm = T),  col='mean gene density'), formula=y~x, method='loess', se=FALSE) +
  geom_smooth(aes(x=start, y=transcripts.length.mean/max(transcripts.length.mean, na.rm = T),  col='mean gene length'), formula=y~x, method='loess', se=FALSE) +
  geom_smooth(aes(x=start, y=operons.length.mean/max(operons.length.mean,na.rm=T),  col='mean operon length'), formula=y~x, method='loess', se=FALSE) +
  geom_smooth(aes(x=start, y=operons.count.mean/max(operons.count.mean,na.rm=T),  col='mean operon density'), formula=y~x, method='loess', se=FALSE) +
  ylab("Normalized mean per tile across all species") +
  facet_wrap(seqnames~Oki.Arm, nrow=1)
p1

p2 <- ggplot( tb ) + aes(x=start, y=transcripts.count.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1, scales = 'free_y') + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('gene density')

p3 <- ggplot( tb ) + aes(x=start, y=transcripts.length.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1, scales = 'free_y') + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('gene size') 

p4 <- ggplot( tb ) + aes(x=start, y=operons.count.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1, scales = 'free_y') + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('operon density') 

p5 <- ggplot( tb ) + aes(x=start, y=operons.length.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1, scales = 'free_y') + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('operon size') 

p6 <- ggplot( tb ) + aes(x=start, y=transcripts.dNdS.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1, scales = 'free_y') + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('dN/dS') 

p7 <- ggplot( tb ) + aes(x=start, y=breakpoints.count.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1, scales = 'free_y') + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('breakpoints') 

p8 <- ggplot( tb ) + aes(x=start, y=repeats.count.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1, scales = 'free_y') + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('repeats') 

p9 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.dNdS.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(~seqnames_arm, nrow=1, scales = 'free') + ylab('dN/dS')

(p2 / p3 / p4 / p5 / p6 / p7 / p8 / p9) # all plots. Too much

(p2 / p3 / p4 / p5 / p6 / p7) # No correlation

(p3 / p5 / p6 / p7) # No densities

(p3 / p5 / p6 / p7 / p8 ) # No densities, repeats

(p3 / p5 / p6 / p7 / p9 ) # No densities + correlation

```

Gene size and density vs. operon size and density, fixed scales.
```{r chromosome_plots_tiles_5_3, dependson="chromosome_plots_tiles_calcs", fig.height=10, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p2 <- ggplot( tb ) + aes(x=start, y=transcripts.count.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('gene density')

p3 <- ggplot( tb ) + aes(x=start, y=transcripts.length.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('gene size')

p4 <- ggplot( tb ) + aes(x=start, y=operons.count.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('operon density')

p5 <- ggplot( tb ) + aes(x=start, y=operons.length.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('operon size')

p6 <- ggplot( tb ) + aes(x=start, y=transcripts.dNdS.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('dN/dS')

p7 <- ggplot( tb ) + aes(x=start, y=breakpoints.count.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('breakpoints')

p8 <- ggplot( tb ) + aes(x=start, y=repeats.count.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('repeats')

p9 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.dNdS.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(~seqnames_arm, nrow=1, scales = 'free') + ylab('dN/dS')

(p2 / p3 / p4 / p5 / p6 / p7 / p8 / p9) # all plots. Too much

(p2 / p3 / p4 / p5 / p6 / p7) # No correlation

(p3 / p5 / p6 / p7) # No densities

(p3 / p5 / p6 / p7 / p8 ) # No densities, repeats

(p3 / p5 / p6 / p7 / p9 ) # No densities + correlation

(p3 / p5 / p6 / p7 / p8 / p9 ) # No densities, repeats, correlation

```

Gene size and density vs. operon size and density, boxplot edition.

```{r chromosome_plots_tiles_5_4, dependson="chromosome_plots_tiles_calcs", fig.height=10, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p2 <- ggplot( tb ) + aes(x=start, y=transcripts.count.mean, col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1) + ylab('gene density')

p3 <- ggplot( tb ) + aes(x=start, y=transcripts.length.mean, col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1)  + ylab('gene size') 

p4 <- ggplot( tb ) + aes(x=start, y=operons.count.mean, col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1) + ylab('operon density') 

p5 <- ggplot( tb ) + aes(x=start, y=operons.length.mean, col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1) + ylab('operon size') 

p6 <- ggplot( tb ) + aes(x=start, y=transcripts.dNdS.mean, col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1) + ylab('dN/dS') 

p7 <- ggplot( tb ) + aes(x=start, y=breakpoints.count.mean, col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1) + ylab('breakpoints') 

p8 <- ggplot( tb ) + aes(x=start, y=repeats.count.mean, col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1) + ylab('repeats') 

p9 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.dNdS.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(~seqnames_arm, nrow=1, scales = 'free') + ylab('dN/dS')

(p2 / p3 / p4 / p5 / p6 / p7 / p8 / p9) # all plots. Too much

(p2 / p3 / p4 / p5 / p6 / p7) # No correlation

(p3 / p5 / p6 / p7) # No densities

(p3 / p5 / p6 / p7 / p8 ) # No densities, repeats

(p3 / p5 / p6 / p7 / p9 ) # No densities + correlation

(p3 / p5 / p6 / p7 / p8 / p9 ) # No densities, repeats, correlation

```

`ggboxplot` version, for statistics.

```{r chromosome_plots_tiles_5_5, dependson="chromosome_plots_tiles_calcs", fig.height=10, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
library(ggpubr)
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p2 <- ggboxplot(tb, x='seqnames_arm', y='transcripts.count.mean', color='Oki.Arm', add='jitter') + theme(axis.text.x=element_text(angle=90)) + ylab('gene density') + stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p3 <- ggboxplot(tb, x='seqnames_arm', y='transcripts.length.mean', color='Oki.Arm', add='jitter') + theme(axis.text.x=element_text(angle=90)) + ylab('gene size') + stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p4 <- ggboxplot(tb, x='seqnames_arm', y='operons.count.mean', color='Oki.Arm', add='jitter') + theme(axis.text.x=element_text(angle=90)) + ylab('operon density') + stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p5 <- ggboxplot(tb, x='seqnames_arm', y='operons.length.mean', color='Oki.Arm', add='jitter') + theme(axis.text.x=element_text(angle=90)) + ylab('operon size') + stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p6 <- ggboxplot(tb, x='seqnames_arm', y='transcripts.dNdS.mean', color='Oki.Arm', add='jitter') + theme(axis.text.x=element_text(angle=90)) + ylab('dN/dS') + stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p7 <- ggboxplot(tb, x='seqnames_arm', y='breakpoints.count.mean', color='Oki.Arm', add='jitter') + theme(axis.text.x=element_text(angle=90)) + ylab('breakpoints') + stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p8 <- ggboxplot(tb, x='seqnames_arm', y='repeats.count.mean', color='Oki.Arm', add='jitter') + theme(axis.text.x=element_text(angle=90)) + ylab('repeats') + stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p9 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.dNdS.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(~seqnames_arm, nrow=1, scales = 'free') + ylab('dN/dS')

(p2 / p3 / p4 / p5 / p6 / p7 / p8 / p9) # all plots. Too much

(p2 / p3 / p4 / p5 / p6 / p7) # No correlation

(p3 / p5 / p6 / p7) # No densities

(p3 / p5 / p6 / p7 / p8 ) # No densities, repeats

(p3 / p5 / p6 / p7 / p9 ) # No densities + correlation

(p3 / p5 / p6 / p7 / p8 / p9 ) # No densities, repeats, correlation

detach('package:ggpubr')

```


```{r chromosome_plots_tiles_6, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
chrt_unified |>
  as.data.frame() |>
  subset(select=c("seqnames", "start", "end", "Oki.Arm", "categories.isolated.mean", "categories.collinear.mean", "categories.breakpoint.mean", "categories.bridge.mean" )) |>
 tidyr::pivot_longer(c("categories.isolated.mean", "categories.collinear.mean", "categories.breakpoint.mean", "categories.bridge.mean")) |>
  ggplot() +
    aes(value, paste0(seqnames, "_", Oki.Arm),  fill = name) +
    geom_bar(position='fill', stat='identity') +
    ylab("chr and arm") +
    ggtitle("Tiles (averages across Oki, Osa, Bar): Alignment categories by location")
```

```{r chromosome_plots_tiles_7, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
library(ggpubr)
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p0 <- tb |> subset(select=c("seqnames", "seqnames_arm", "start", "end", "Oki.Arm", "categories.isolated.mean", "categories.collinear.mean", "categories.breakpoint.mean", "categories.bridge.mean" )) |> tidyr::pivot_longer(c("categories.isolated.mean", "categories.collinear.mean", "categories.breakpoint.mean", "categories.bridge.mean")) |>
   ggplot() +
    aes(value, seqnames_arm,  fill = name) +
    facet_wrap(seqnames~Oki.Arm, nrow=1, scales='free') +
    geom_bar(position='fill', stat='identity') +
    theme(axis.text.y=element_blank(), axis.ticks.y =element_blank()) +
    ylab("chr and arm") +
    xlab("category")
p0 + ggtitle("Tiles (averages across Oki, Osa, Bar): Alignment categories across genome")

p1 <- ggplot( tb ) + aes(x=start, y=transcripts.dNdS.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('dN/dS') 
p1 + ggtitle("Tiles (averages across Oki, Osa, Bar): Properties across genome")

p2 <- ggplot( tb ) + aes(x=start, y=breakpoints.count.mean , col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess', method.args = list(family = "symmetric")) + ylab('breakpoints') 
p2 + ggtitle("Tiles (averages across Oki, Osa, Bar): Properties across genome")

p3 <- ggboxplot(tb, x='seqnames_arm', y='breakpoints.count.mean', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('breakpoints') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') +
  theme_classic()
p3 + ggtitle('Tiles (averages across Oki, Osa, Bar): Breakpoints per chr+arm')

p4 <- ggboxplot(tb, x='seqnames_arm', y='transcripts.dNdS.mean', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('dN/dS') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') +
  theme_classic()
p4 +  ggtitle('Tiles (averages across Oki, Osa, Bar): dNdS per chr+arm')

# And a correlation plot
p5 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.dNdS.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('dN/dS')

p5 + ggtitle('Tiles (averages across Oki, Osa, Bar): dNdS per chr+arm') 

p6 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=repeats.count.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('repeats')

p6 

layout <- "
A
B
C
D
E
F
"

( p0 / p1 / p4 / p2 / p3 / p5 ) #+ plot_layout(design = layout, guides = 'collect')
( p0 / p1 / p4 / p2 / p3 )
( p1 / p2 / p3 / p4 )
( p1 / p2 / p3 )
( p0 / p1 / p2 / p3 )

detach('package:ggpubr')
```


```{r chromosome_plots_tiles_7_1, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
library(ggpubr)
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)
#tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Osa.Arm)
#tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Bar.Arm)

p0 <- tb |> subset(select=c("seqnames", "seqnames_arm", "start", "end", "Oki.Arm", "categories.isolated.mean", "categories.collinear.mean", "categories.breakpoint.mean", "categories.bridge.mean" )) |> tidyr::pivot_longer(c("categories.isolated.mean", "categories.collinear.mean", "categories.breakpoint.mean", "categories.bridge.mean")) |>
   ggplot() +
    aes(value, seqnames_arm,  fill = name) +
    facet_wrap(seqnames~Oki.Arm, nrow=1, scales='free') +
    geom_bar(position='fill', stat='identity') +
    theme(axis.text.y=element_blank(), axis.ticks.y =element_blank()) +
    ylab("chr and arm") +
    xlab("category")

p0 + ggtitle("Tiles (averages across Oki, Osa, Bar): Alignment categories by position")

p1 <- ggplot(tb) + aes(x=start, y=transcripts.dNdS.mean, col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1) + ylab('dN/dS')
p1 + ggtitle("Tiles (averages across Oki, Osa, Bar): dN/dS across genome")
p1 <- p1 + theme(axis.title.x=element_blank()) # remove X labels for later

p2 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=breakpoints.count.mean , col = Oki.Arm) + geom_boxplot(outlier.shape=NA) + geom_point() + facet_wrap(~seqnames, nrow = 1) + ylab('Breakpoints')
p2 + ggtitle("Tiles (averages across Oki, Osa, Bar): Breakpoints across genome")

# And a correlation plot
p3 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.dNdS.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('dN/dS')

p3 + ggtitle('Tiles (averages across Oki, Osa, Bar): dNdS per chr+arm')

p3 <- p3 + theme(axis.title.x=element_blank()) # remove X labels for later

p4 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=repeats.count.mean) + geom_point() + geom_smooth(method='lm', col='red' , formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('repeats') + xlab('Breakpoints')
p4

( p0 / p1 / p2 / p3 / p4 ) #+ plot_layout(design = layout, guides = 'collect')
( p1 / p2 / p3 / p4 )
( p1 / p2 / p3 )
( p0 / p1 / p2 / p3 )

detach('package:ggpubr')
```

## Checking my work

Double-checking breakpoint patterns to make sure they make sense and match the averages.

```{r chromosome_plots_tiles_8_bp_symmetric, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- chrt_unified |> as.data.frame() |> subset(select=c("seqnames", "start", "end", "Oki.bp.Oki_Osa.breaks.count", "Oki.bp.Oki_Osa.breaks.count", "Oki.bp.Oki_Bar.breaks.count", "Oki.bp.Oki_Kum.breaks.count", "Oki.bp.Oki_Aom.breaks.count", "Oki.bp.Oki_Nor.breaks.count")) |> tidyr::pivot_longer(c("Oki.bp.Oki_Osa.breaks.count", "Oki.bp.Oki_Osa.breaks.count", "Oki.bp.Oki_Bar.breaks.count", "Oki.bp.Oki_Kum.breaks.count", "Oki.bp.Oki_Aom.breaks.count", "Oki.bp.Oki_Nor.breaks.count")) |> ggplot() + aes(start, value, col=name) + facet_wrap(~seqnames, nrow=1) + geom_smooth(method='loess', formula=y~x, span=0.25, se=FALSE, method.args=list(family="symmetric"))

p2 <- chrt_unified |> as.data.frame() |> subset(select=c("seqnames", "start", "end", "Osa.bp.Osa_Oki.breaks.count", "Osa.bp.Osa_Bar.breaks.count", "Osa.bp.Osa_Aom.breaks.count", "Osa.bp.Osa_Kum.breaks.count", "Osa.bp.Osa_Nor.breaks.count")) |> tidyr::pivot_longer(c("Osa.bp.Osa_Oki.breaks.count", "Osa.bp.Osa_Bar.breaks.count", "Osa.bp.Osa_Aom.breaks.count", "Osa.bp.Osa_Kum.breaks.count", "Osa.bp.Osa_Nor.breaks.count")) |> ggplot() + aes(start, value, col=name) + facet_wrap(~seqnames, nrow=1) + geom_smooth(method='loess', formula=y~x, span=0.25, se=FALSE, method.args=list(family="symmetric"))

p3 <- chrt_unified |> as.data.frame() |> subset(select=c("seqnames", "start", "end", "Bar.bp.Bar_Oki.breaks.count", "Bar.bp.Bar_Osa.breaks.count", "Bar.bp.Bar_Aom.breaks.count", "Bar.bp.Bar_Kum.breaks.count", "Bar.bp.Bar_Nor.breaks.count")) |> tidyr::pivot_longer(c("Bar.bp.Bar_Oki.breaks.count", "Bar.bp.Bar_Osa.breaks.count", "Bar.bp.Bar_Aom.breaks.count", "Bar.bp.Bar_Kum.breaks.count", "Bar.bp.Bar_Nor.breaks.count")) |> ggplot() + aes(start, value, col=name) + facet_wrap(~seqnames, nrow=1) + geom_smooth(method='loess', formula=y~x, span=0.25, se=FALSE, method.args=list(family="symmetric"))

(p1 / p2 / p3)
```

Same with dN/dS.

```{r chromosome_plots_tiles_8_dnds_symmetric, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- chrt_unified |> as.data.frame() |> ggplot() + aes(start, Oki.transcripts.dNdS_GUIDANCE2.median, col=Oki.Arm) + facet_wrap(~seqnames, nrow=1) + geom_point() + geom_smooth(method='loess', formula=y~x, span=0.25, se=T, method.args=list(family="symmetric"))
p2 <- chrt_unified |> as.data.frame() |> ggplot() + aes(start, Osa.transcripts.dNdS_GUIDANCE2.median, col=Osa.Arm) + facet_wrap(~seqnames, nrow=1) + geom_point() + geom_smooth(method='loess', formula=y~x, span=0.25, se=T, method.args=list(family="symmetric"))
p3 <- chrt_unified |> as.data.frame() |> ggplot() + aes(start, Bar.transcripts.dNdS_GUIDANCE2.median, col=Bar.Arm) + facet_wrap(~seqnames, nrow=1) + geom_point() + geom_smooth(method='loess', formula=y~x, span=0.25, se=T, method.args=list(family="symmetric"))
(p1 / p2 / p3)
```

Double-check the smoothing.

```{r chromosome_plots_tiles_8_bp_nonsymmetric, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- chrt_unified |> as.data.frame() |> subset(select=c("seqnames", "start", "end", "Oki.bp.Oki_Osa.breaks.count", "Oki.bp.Oki_Osa.breaks.count", "Oki.bp.Oki_Bar.breaks.count", "Oki.bp.Oki_Kum.breaks.count", "Oki.bp.Oki_Aom.breaks.count", "Oki.bp.Oki_Nor.breaks.count")) |> tidyr::pivot_longer(c("Oki.bp.Oki_Osa.breaks.count", "Oki.bp.Oki_Osa.breaks.count", "Oki.bp.Oki_Bar.breaks.count", "Oki.bp.Oki_Kum.breaks.count", "Oki.bp.Oki_Aom.breaks.count", "Oki.bp.Oki_Nor.breaks.count")) |> ggplot() + aes(start, value, col=name) + facet_wrap(~seqnames, nrow=1) + geom_smooth(method='loess', formula=y~x, span=0.25, se=FALSE)

p2 <- chrt_unified |> as.data.frame() |> subset(select=c("seqnames", "start", "end", "Osa.bp.Osa_Oki.breaks.count", "Osa.bp.Osa_Bar.breaks.count", "Osa.bp.Osa_Aom.breaks.count", "Osa.bp.Osa_Kum.breaks.count", "Osa.bp.Osa_Nor.breaks.count")) |> tidyr::pivot_longer(c("Osa.bp.Osa_Oki.breaks.count", "Osa.bp.Osa_Bar.breaks.count", "Osa.bp.Osa_Aom.breaks.count", "Osa.bp.Osa_Kum.breaks.count", "Osa.bp.Osa_Nor.breaks.count")) |> ggplot() + aes(start, value, col=name) + facet_wrap(~seqnames, nrow=1) + geom_smooth(method='loess', formula=y~x, span=0.25, se=FALSE)

p3 <- chrt_unified |> as.data.frame() |> subset(select=c("seqnames", "start", "end", "Bar.bp.Bar_Oki.breaks.count", "Bar.bp.Bar_Osa.breaks.count", "Bar.bp.Bar_Aom.breaks.count", "Bar.bp.Bar_Kum.breaks.count", "Bar.bp.Bar_Nor.breaks.count")) |> tidyr::pivot_longer(c("Bar.bp.Bar_Oki.breaks.count", "Bar.bp.Bar_Osa.breaks.count", "Bar.bp.Bar_Aom.breaks.count", "Bar.bp.Bar_Kum.breaks.count", "Bar.bp.Bar_Nor.breaks.count")) |> ggplot() + aes(start, value, col=name) + facet_wrap(~seqnames, nrow=1) + geom_smooth(method='loess', formula=y~x, span=0.25, se=FALSE)

(p1 / p2 / p3)
```

```{r chromosome_plots_tiles_8_dnds_nonsymmetric, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- chrt_unified |> as.data.frame() |> ggplot() + aes(start, Oki.transcripts.dNdS_GUIDANCE2.median, col=Oki.Arm) + facet_wrap(~seqnames, nrow=1) + geom_point() + geom_smooth(method='loess', formula=y~x, span=0.25, se=T)

p2 <- chrt_unified |> as.data.frame() |> ggplot() + aes(start, Osa.transcripts.dNdS_GUIDANCE2.median, col=Osa.Arm) + facet_wrap(~seqnames, nrow=1) + geom_point() + geom_smooth(method='loess', formula=y~x, span=0.25, se=T)

p3 <- chrt_unified |> as.data.frame() |> ggplot() + aes(start, Bar.transcripts.dNdS_GUIDANCE2.median, col=Bar.Arm) + facet_wrap(~seqnames, nrow=1) + geom_point() + geom_smooth(method='loess', formula=y~x, span=0.25, se=T)

(p1 / p2 / p3)
```

The same figure as above, but without symmetric smoothing.

```{r chromosome_plots_tiles_9, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=transcripts.dNdS.mean, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('dN/dS')

p2 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=breakpoints.count.mean , col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('breakpoints')

library(ggpubr)
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p3 <- ggboxplot(tb, x='seqnames_arm', y='breakpoints.count.mean', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('breakpoints') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p4 <- ggboxplot(tb, x='seqnames_arm', y='transcripts.dNdS.mean', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('dN/dS') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p5 <- ggplot(tb) + aes(x=breakpoints.count.mean, y=transcripts.dNdS.mean) + geom_point() + geom_smooth(method='lm', col='red', formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('dN/dS')

layout <- "
A
B
C
D
E
"

( p1 / p4 / p2 / p3 / p5 ) + plot_layout(design = layout, guides = 'collect')

detach('package:ggpubr')
```

Finally, double check the look of the figure with different pairs of species.

Start with Okinawa-Osaka.

```{r chromosome_plots_tiles_9_oki_osa, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=Oki.transcripts.dNdS_GUIDANCE2.median, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('dN/dS')

p2 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=Oki.bp.Oki_Osa.breaks.count , col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('breakpoints')

library(ggpubr)
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p3 <- ggboxplot(tb, x='seqnames_arm', y='Oki.bp.Oki_Osa.breaks.count', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('breakpoints') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p4 <- ggboxplot(tb, x='seqnames_arm', y='Oki.transcripts.dNdS_GUIDANCE2.median', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('dN/dS') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p5 <- ggplot(tb) + aes(x=Oki.bp.Oki_Osa.breaks.count, y=Oki.transcripts.dNdS_GUIDANCE2.median) + geom_point() + geom_smooth(method='lm', col='red', formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('dN/dS')

layout <- "
A
B
C
D
E
"

( p1 / p4 / p2 / p3 / p5 ) + plot_layout(design = layout, guides = 'collect')

detach('package:ggpubr')
```

The Okinawa-Barcelona...

```{r chromosome_plots_tiles_9_oki_bar, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=Oki.transcripts.dNdS_GUIDANCE2.median, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('dN/dS')

p2 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=Oki.bp.Oki_Bar.breaks.count , col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('breakpoints')

library(ggpubr)
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p3 <- ggboxplot(tb, x='seqnames_arm', y='Oki.bp.Oki_Bar.breaks.count', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('breakpoints') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p4 <- ggboxplot(tb, x='seqnames_arm', y='Oki.transcripts.dNdS_GUIDANCE2.median', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('dN/dS') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p5 <- ggplot(tb) + aes(x=Oki.bp.Oki_Bar.breaks.count, y=Oki.transcripts.dNdS_GUIDANCE2.median) + geom_point() + geom_smooth(method='lm', col='red', formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('dN/dS')

layout <- "
A
B
C
D
E
"

( p1 / p4 / p2 / p3 / p5 ) + plot_layout(design = layout, guides = 'collect')

detach('package:ggpubr')
```

Then Osaka-Barcelona...

```{r chromosome_plots_tiles_9_osa_bar, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=Osa.transcripts.dNdS_GUIDANCE2.median, col = Osa.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('dN/dS')

p2 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=Osa.bp.Osa_Bar.breaks.count , col = Osa.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('breakpoints')

library(ggpubr)
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Osa.Arm)

p3 <- ggboxplot(tb, x='seqnames_arm', y='Osa.bp.Osa_Bar.breaks.count', color='Osa.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('breakpoints') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p4 <- ggboxplot(tb, x='seqnames_arm', y='Osa.transcripts.dNdS_GUIDANCE2.median', color='Osa.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('dN/dS') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p5 <- ggplot(tb) + aes(x=Osa.bp.Osa_Bar.breaks.count, y=Osa.transcripts.dNdS_GUIDANCE2.median) + geom_point() + geom_smooth(method='lm', col='red', formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Osa.Arm, nrow=1, scales = 'free') + ylab('dN/dS')

layout <- "
A
B
C
D
E
"

( p1 / p4 / p2 / p3 / p5 ) + plot_layout(design = layout, guides = 'collect')

detach('package:ggpubr')
```

And finally Okinawa-Kume.

```{r chromosome_plots_tiles_9_oki_kum, dependson="chromosome_plots_tiles_calcs", fig.height=6, fig.width=12, message=FALSE, warning=FALSE, dev=c("svg", "png")}
p1 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=Oki.transcripts.dNdS_GUIDANCE2.median, col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('dN/dS')

p2 <- ggplot( chrt_unified |> as.data.frame() ) + aes(x=start, y=Oki.bp.Oki_Kum.breaks.count , col = Oki.Arm) + geom_point() + facet_wrap(~seqnames, nrow = 1) + geom_smooth(formula=y~x, method = 'loess') + ylab('breakpoints')

library(ggpubr)
tb <- chrt_unified |> as.data.frame()
tb$seqnames_arm <- paste0(tb$seqnames, "_", tb$Oki.Arm)

p3 <- ggboxplot(tb, x='seqnames_arm', y='Oki.bp.Oki_Kum.breaks.count', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('breakpoints') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p4 <- ggboxplot(tb, x='seqnames_arm', y='Oki.transcripts.dNdS_GUIDANCE2.median', color='Oki.Arm', add='jitter') +
  theme(axis.text.x=element_text(angle=90)) +
  ylab('dN/dS') +
  stat_compare_means(comparison=list(c('Chr1_short', 'Chr1_long'), c('Chr2_short', 'Chr2_long'), c('PAR_short', 'PAR_long'), c('XSR_XSR', 'PAR_short'), c('XSR_XSR', 'PAR_long')), paired = FALSE, na.rm = TRUE, method='t.test') + theme_classic()

p5 <- ggplot(tb) + aes(x=Oki.bp.Oki_Kum.breaks.count, y=Oki.transcripts.dNdS_GUIDANCE2.median) + geom_point() + geom_smooth(method='lm', col='red', formula=y~x) + ggpubr::stat_cor() + facet_wrap(seqnames~Oki.Arm, nrow=1, scales = 'free') + ylab('dN/dS')

layout <- "
A
B
C
D
E
"

( p1 / p4 / p2 / p3 / p5 ) + plot_layout(design = layout, guides = 'collect')

#detach('package:ggpubr')
```




# Session information

```{r sessionInfo}
sessionInfo()
```
