---
title: "Distance Measures"
author: 
 - "Charles Plessy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Distance Measures}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

# Purpose

We want to put numbers on visual impressions such as:

 - There is a lot of scrambling.
 - Aligned regions stay on their chromosome arm.
 - Aligned regions are more "off the diagonal" on short arms than on long arms.
 - …

# Load R pacakges and data

```{r load_packages_and_data}
library('OikScrambling') |> suppressPackageStartupMessages()
load("BreakPoints.Rdata")
```

See `vignette("LoadGenomicBreaks", package = "OikScrambling")` for how the
different GBreaks objects are prepared.

# Synteny index

Ad-hoc index measuring to what extent a scaffold of the _target_ genome is
mostly aligned to a single scaffold in the _query_ genome.   See
`?GenomicBreaks::synteny_index` for details.  One limitation to the use of this
index is that it requires that at least the _query_ genome is a complete
chromosome assembly.

```{r synteny_index}
pairs2table6x3 <- function (x) {
  x <- as.list(x)
  cbind(
    Oki = c(Oki = NA,        Osa = x$Osa_Oki, Bar = x$Bar_Oki),
    Kum = c(Oki = x$Oki_Kum, Osa = x$Osa_Kum, Bar = x$Bar_Kum),
    Osa = c(Oki = x$Oki_Osa, Osa = NA,        Bar = x$Bar_Osa),
    Aom = c(Oki = x$Oki_Aom, Osa = x$Osa_Aom, Bar = x$Bar_Aom),
    Bar = c(Oki = x$Oki_Bar, Osa = x$Osa_Bar, Bar = NA       ),
    Nor = c(Oki = x$Oki_Nor, Osa = x$Osa_Nor, Bar = x$Bar_Nor)
  )
}

pairs2Kable6x3 <- function (df) df |> pairs2table6x3() |> round(2) |> knitr::kable()

BiocParallel::bplapply(gbs,                 synteny_index  ) |> pairs2Kable6x3()
BiocParallel::bplapply(gbs, \(g) swap(g) |> synteny_index()) |> pairs2Kable6x3()
```

# Correlation

Ad-hoc index measuring the correlation of the coordinates of the syntenic
alignments in scaffolds of a _target_ genome and their best match in the
_query_ genome.  See `?GenomicBreaks::correlation_index` for details.  It is
intended to be more robust to the presence of uncollapsed haplotypes in the
_query_ genome but probably needs further testing.

```{r correlation_index}
BiocParallel::bplapply(gbs,                 correlation_index  ) |> pairs2Kable6x3()
BiocParallel::bplapply(gbs, \(g) swap(g) |> correlation_index()) |> pairs2Kable6x3()
```

# Gene Order Conservation

See `?GenomicBreaks::GOC` for details.

```{r goc_index}
BiocParallel::bplapply(gbs,                 GOC  ) |> pairs2Kable6x3()
BiocParallel::bplapply(gbs, \(g) swap(g) |> GOC()) |> pairs2Kable6x3()
```

# Strand randomisation index

See `?GenomicBreaks::strand_randomisation_index()` for details.

```{r strand_randomisation_index}
BiocParallel::bplapply(gbs,                 strand_randomisation_index  ) |> pairs2Kable6x3()
BiocParallel::bplapply(gbs, \(g) swap(g) |> strand_randomisation_index()) |> pairs2Kable6x3()
```
