---
title: "Introduction to the OikScrambling package"
author: 
 - "Charlotte West"
 - "Charles Plessy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Introduction to the OikScrambling package}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE)
knitr::opts_knit$set(verbose = TRUE)
```

Load packages and data
======================

See `?OikScrambling:::loadAllGenomes`, `?OikScrambling:::loadAllTranscriptsGR`,
and `vignette("LoadGenomicBreaks", package = "OikScrambling")` for how the
different objects are prepared.

```{r load_packages_and_data}
library('OikScrambling')   |> suppressPackageStartupMessages()
library('GenomicFeatures') |> suppressPackageStartupMessages()
(genomes     <- OikScrambling:::loadAllGenomes())
(transcripts <- OikScrambling:::loadAllTranscriptsGR() |> suppressWarnings())
(annots      <- OikScrambling:::loadAllAnnotations()   |> suppressWarnings())
(load("BreakPoints.Rdata"))
requireNamespace("rGADEM")
requireNamespace("ggseqlogo")
```

# Tandem repeats

Repeated regions of genomes are notoriously difficult to assemble and align. Thus, information about an alignment stop's proximity to a tandem repeat is useful information in breakpoint classification. Given a user-defined tolerance, the function `tan_bp` will classify alignment stops as either far from, close to, or within tandem repeats (elements 1,2 and 3 of the produced list, respectively). The function requires the tandem repeat coverage of one of the genomes. By default, the function will expect the target genome tandem coverage. Setting `query_tf = TRUE` will switch the outcome to be query genome oriented.

```{r echo=T, message=F, warning=F}
# 
# tan_O <- rtracklayer::import.gff3(system.file("extdata/OSKA2016.tantanf4w100.gff3.gz",  package="GenomicBreaks"))
# tan_O <- GRanges(tan_O, seqinfo = seqinfo(OSKA2016))
# tan_gr_O <- tan_bp(gr_ob = gr_O_Oki, tan = tan_O, tol = 50)
# tan_gr_O
# 
# tan_Oki <- rtracklayer::import.gff3(system.file("extdata/I69-5.tantanf4w100.gff3.gz",  package="GenomicBreaks"))
# tan_Oki <- GRanges(tan_Oki, seqinfo = seqinfo(OKI2018_I69))
# tan_gr_Oki <- tan_bp(gr_ob = gr_O_Oki, tan = tan_Oki, tol = 50, query_tf = TRUE)
# tan_gr_Oki

```

We can use another script to plot the density of tandem repeats around alignment stops (and breakpoints). In the following plots, breakpoints are aligned at the centre of the plots. Intuitively, this function can be generalised to see the density of any binary coverage characteristic around breakpoints, given a GRanges object with per-base coverage of a given characteristic. Here we stick with tandem repeat info:

```{r echo=T, message=F, warning=F, cache=FALSE}
# hm_tan_O <- gr_O_Oki %>%
#   feature_coverage(tan_O, win = 1000, lab = "Osaka ~ Okinawa", direction = "left") %>%
#   smoothHeatmap(output.size = c(2000,500), algorithm = "kernel") %T>%
#   plotHeatmap
# 
# hm_red_tan_O <- reduced_gr_O_Oki %>%
#   feature_coverage(tan_O, win = 1000, lab = "Osaka ~ Okinawa", direction = "left") %>%
#   smoothHeatmap(output.size = c(2000,500), algorithm = "kernel") %T>%
#   plotHeatmap
# 
# hm_tan_Oki <- gr_Oki_O %>%
#   feature_coverage(tan_Oki, win = 1000, lab = "Osaka ~ Okinawa", direction = "left") %>%
#   smoothHeatmap(output.size = c(2000,500), algorithm = "kernel") %T>%
#   plotHeatmapList
```

# Coverage

Another characteristic of alignment stops that we can examine is the coverage depth over said alignment stops. By aligning raw reads to assembled genomes, we can obtain per-base coverage depth for target or query genome. Low coverage directly on and surrounding an alignment stop may suggest unreliability in it being a true breakpoint. The function `bp_coverage` returns a GRanges object of the bps, with associated averaged and point converage in the metadatacolumns. 

```{r echo=T, message=F, warning=F}

# cov_gr_O_Oki <- bp_coverage(gr_ob = q_Oki, cov_gr = Oki_cov_pb, win = 50)
# cov_gr_O_Oki

```



# Nucleic acid content heatmaps

The `GenomicBreaks::bp_heatmap()` function plots nucleic acid content in windows
centered on alignment stops.  Lets look at `GC` content before and after
coalescing: 

```{r echo=T, fig.height=6, fig.width=12, message=FALSE, warning=FALSE}
library(heatmaps)
#hm_GC1 <- sapply(gbs,  bp_heatmap, basep_range = 1000, pat = "GC", direction = "left") |> SimpleList()
hm_GC2 <- sapply(coa[1:15],  bp_heatmap, basep_range = 1000, pat = "GC", direction = "left") |> SimpleList()
#hm_GC3 <- sapply(coa2, bp_heatmap, basep_range = 1000, pat = "GC", direction = "left") |> SimpleList()

maxHeight <- 2000
#sm_GC1 <- sapply(hm_GC1, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()
sm_GC2 <- sapply(hm_GC2, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()
#sm_GC3 <- sapply(hm_GC3, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()

#plotHeatmapList(sm_GC1[1:5])

plotHeatmapList(sm_GC2[1:5])
#plotHeatmapList(sm_GC2[6:10])
#plotHeatmapList(sm_GC2[11:15])

#plotHeatmapList(sm_GC3[1:5])
```

The alignment stops are ordered in such a way that start (or left breaks) are
centred on the top half of the plot, and end alignment stops are centred on the
bottom. This is why we see different directionality in the plots. At the
alignment stops, the GC content seems to be lower, but higher once within the
aligned region. 

Now looking at TATA content:

```{r TATA, fig.height=6, fig.width=12}
#hm_GC1_TATA <- sapply(gbs, bp_heatmap, basep_range = 1000, pat = "TATA", direction = "left") |> SimpleList()
hm_GC2_TATA <- sapply(coa[1:15], bp_heatmap, basep_range = 1000, pat = "TATA", direction = "left") |> SimpleList()

maxHeight <- 2000
#sm_GC1_TATA <- sapply(hm_GC1_TATA, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()
sm_GC2_TATA <- sapply(hm_GC2_TATA, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()
#sm_GC3_TATA <- sapply(hm_GC3_TATA, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()

#plotHeatmapList(sm_GC1_TATA)
plotHeatmapList(sm_GC2_TATA[1:5])
#plotHeatmapList(sm_GC3_TATA)
```

The directionaility here is consistent with the `GC` content analysis.
Furthermore; "TATA box sequence can act as a basal promoter element not only for
RNA polymerase II (RNAP II) transcription, but also for transcription by RNA
polymerase III (RNAP III)" - Wang Y, Jensen RC, Stumph WE. Role of TATA box
sequence and orientation in determining RNA polymerase II/III transcription
specificity. Nucleic Acids Res. 1996;24(15):3100–3106.
doi:10.1093/nar/24.15.3100. However, the areas are not particularly enriched
for TATA boxes, so it does not necessarily show that breakpoints are occurring
directly after promoter regions. 


```{r AAGCsGCwwmkCGrCTTyn, fig.height=6, fig.width=12}
#hm_GC1_mot <- sapply(gbs, bp_heatmap, basep_range = 1000, pat = "AGCNGC", direction = "left") |> SimpleList()
hm_GC2_mot <- sapply(coa[1:15], bp_heatmap, basep_range = 1000, pat = "AGCNGC", direction = "left") |> SimpleList()

maxHeight <- 2000
#sm_GC1_mot <- sapply(hm_GC1_mot, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()
sm_GC2_mot <- sapply(hm_GC2_mot, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()
#sm_GC3_mot <- sapply(hm_GC3_mot, smoothHeatmap, output.size = c(maxHeight,     500), algorithm = "kernel") |> SimpleList()

#plotHeatmapList(sm_GC1_mot)
plotHeatmapList(sm_GC2_mot[1:5])
#plotHeatmapList(sm_GC3_mot)
```


## Cluster analysis


```{r cluster_analysis, fig.height=20, fig.width=20}
# # Accessory functions
# clusterHeatmap <- function(hm, k = 2) kmeans(image(hm), k)
# 
# orderHeatmap <- function(hm, cl) {
#   image(hm) <- image(hm)[order(cl$cluster),]
#   hm
# }
# 
# subsetHeatmap <- function(hm, bool) {
#   image(hm) <- image(hm)[bool,]
#   hm
# }
# 
# # DAta
# hm_GC   <- bp_heatmap(      gr_Oki_O, basep_range  = 1000, pat = "GC",     direction = "left")
# hm_TATA <- bp_heatmap(      gr_Oki_O, basep_range  = 1000, pat = "TATA",   direction = "left")
# hm_tan  <- feature_coverage(gr_Oki_O, tan_Oki, win = 1000, lab = "Tandem", direction = "left")
# 
# # Define 3 K-mean clusters
# clusterHeatmap(hm_GC, 3) -> hm_cl_k4
# 
# # Order and smooth
# shm_GC <- orderHeatmap(hm_GC, hm_cl_k4) %>%
#   smoothHeatmap(output.size = c(2000,500), algorithm = "kernel")
# 
# shm_TATA <- orderHeatmap(hm_TATA, hm_cl_k4) %>%
#   smoothHeatmap(output.size = c(2000,500), algorithm = "kernel")
# 
# shm_tan <- orderHeatmap(hm_tan, hm_cl_k4) %>%
#   smoothHeatmap(output.size = c(2000,500), algorithm = "kernel")
# 
# # Plot the ordered data
# plotHeatmapList(list(shm_GC, shm_TATA, shm_tan))
# par(mfrow = c(1,1)) # plotHeatmapList messes with mrfrow...
# plotHeatmapMeta(list(hm_GC, hm_TATA, hm_tan))
# 
# plotHeatmapMeta(list(subsetHeatmap(hm_GC, hm_cl_k4$cluster == 1),
#                      subsetHeatmap(hm_GC, hm_cl_k4$cluster == 2),
#                      subsetHeatmap(hm_GC, hm_cl_k4$cluster == 3)))
# 
# kmeanOrderedPlot <- function (hm, k=3, seed = 1) {
#   set.seed(seed)
#   
#   cl <- kmeans(image(hm), k)
#   mat <- image(hm)[order(cl$cluster),]
#   
#   kmhm = Heatmap(
#     mat,
#     coords=c(-500,500),
#     label="kmeans",
#     scale=range(mat))
#   
#   plotHeatmapList(kmhm,
#                   cex.label=1.5,
#                   partition=as.vector(table(cl$cluster)),
#                   partition.legend=TRUE,
#                   partition.lines=TRUE,
#                   legend=TRUE,
#                   legend.pos="r",
#                   legend.width=0.3)
# }
# 
# kmeanOrderedPlot(shm_GC)
```

## Evidence for breakpoint hotspots

The function `bp_pair_analysis` takes two pairwise alignments with the same target genome, and plots the alignment stops on to the centred alignment stops of the other. In the plot below, Osaka is there consistent target genome, and as such acts as a coordinate system to relate alignment stops across genomes. Lets produce two of these 3-way analyses; one between Osaka-Okinawa-Norway, and one between Osaka-Okinawa-Aomori

```{r echo=T, message=F, warning=F}

# hm1 <- bp_pair_analysis(gr_ref_q1 = gr_O_Oki, gr_ref_q2 = gr_O_N, win = 1000, lab = "Oki~Nor")
# plotHeatmapMeta(hm1)
# 
# hm2 <- bp_pair_analysis(gr_ref_q1 = gr_O_Oki, gr_ref_q2 = gr_O_A, win = 1000, lab = "Oki~Aom")
# plotHeatmapMeta(hm2)

```

The accumulation of alignment breaks of one pairwise alignment onto another suggests the existence of breakpoint hotspots; regions where breaks are far more likely to occur. Intuitively, this would lead to the presumption of synteny blocks, also. The pattern is consistent after coalescing, too;

```{r echo=T, message=F, warning=F}

# reduced_gr_O_N <- coalesce_contigs(gr_ob = gr_O_N, tol = 500)
# 
# hm3 <- bp_pair_analysis(gr_ref_q1 = reduced_gr_O_Oki, gr_ref_q2 = reduced_gr_O_N, win = 1000, lab = "Oki~Nor")
# plotHeatmapMeta(hm3)

```

## Tandem repeat coverage around alignment stops (breakpoints)

The function `GenomicBreaks::feature_coverage()` will plot the coverage of tandem repeats around
centred alignment stops. 

```{r echo=T, message=F, warning=F, cache=FALSE}
# 
# hm_tan1 <- feature_coverage(gr_ob = gr_O_Oki, feat = tan_O, win = 4000, lab = "Osk ~ Oki", direction = "left")
# plotHeatmapMeta(hm_tan1)
#  
# hm_tan2 <- feature_coverage(gr_ob = reduced_gr_O_Oki, feat = tan_O, win = 4000, lab = "OSK~Oki (coalesced)", direction = "left")
# plotHeatmapMeta(hm_tan2)
# 
# # should I also add comparison of proportions of tandem proximities before and after coalescion?
```

There appears to be less tandem repeats around alignment stops after coalescion,
inferring that perhaps that some are artefacts of difficult alignment near to
tandem repeats. 


## Gene feature coverage around alignment stops (breakpoints)

TODO: analyse separately the breakpoint regions that are in colinear regions?

The function `feature_coverage` will plot the coverage of tandem repeats around
centred alignment stops. 

```{r genomic_features, cache=FALSE}
#current_GB <- gr_Oki_O # To ease recomputing with another object
current_GB <- gbs$Oki_Osa
#current_GB <- coa$Oki_Osa
gadems.invLeftGap <- readRDS("gadems.invLeftGap.Rda")
pwm <- gadems.invLeftGap$Oki_Osa[[1]]
pwm@consensus
pwmHits <- matchPWM(pwm = rGADEM::getPWM(pwm), genomes$Oki)

hmList_OKI <- list()

hmList_OKI$genes      <- feature_coverage(current_GB, genes(annots$Oki),                         win = 2000, lab = "Oki genes", direction = "left")
hmList_OKI$promoters  <- feature_coverage(current_GB, promoters(annots$Oki),                     win = 2000, lab = "Oki promoters", direction = "left")
hmList_OKI$exons      <- feature_coverage(current_GB, exonicParts(annots$Oki),                   win = 2000, lab = "Oki exons", direction = "left")
hmList_OKI$introns    <- feature_coverage(current_GB, intronicParts(annots$Oki),                 win = 2000, lab = "Oki introns", direction = "left")
hmList_OKI$cds        <- feature_coverage(current_GB, cds(annots$Oki),                           win = 2000, lab = "Oki cds", direction = "left")
hmList_OKI$utrs5      <- feature_coverage(current_GB, unlist(fiveUTRsByTranscript(annots$Oki)),  win = 2000, lab = "Oki 5′ UTRs", direction = "left")
hmList_OKI$utrs3      <- feature_coverage(current_GB, unlist(threeUTRsByTranscript(annots$Oki)), win = 2000, lab = "Oki 3′ UTRs", direction = "left")
hmList_OKI$pwm        <- feature_coverage(current_GB, pwmHits                                  , win = 2000, lab = "PWM", direction = "left")

plotHeatmapMeta(hmList_OKI)
plotHeatmapMeta(hmList_OKI$genes)
plotHeatmapMeta(hmList_OKI[c("exons", "cds")])
plotHeatmapMeta(hmList_OKI[c("exons", "introns", "utrs5", "utrs3", "pwm")])
plotHeatmapMeta(hmList_OKI[c("introns", "utrs5", "utrs3", "pwm")])
plotHeatmapMeta(hmList_OKI[c("utrs5", "utrs3", "pwm")])
```

Aligned regions are enriched in exons and depleted in introns.  Is it because
exons are more alignable or is it because it is deleterous to break genes
in exons ?

## Gene feature coverage around PWM hits

Let's look at coverage around PWM hits of the AAGCsGCwwmkCGrCTTyn motif

```{r genomic_features_pwmHits, cache=FALSE}
current_GB <- pwmHits

hmList_PWM_OKI <- list()
hmList_PWM_OKI$genes     <- feature_coverage(current_GB, genes(annots$Oki),                         win = 2000, lab = "Oki genes", direction = "left")
hmList_PWM_OKI$promoters <- feature_coverage(current_GB, promoters(annots$Oki),                     win = 2000, lab = "Oki promoters", direction = "left")
hmList_PWM_OKI$exons     <- feature_coverage(current_GB, exonicParts(annots$Oki),                   win = 2000, lab = "Oki exons", direction = "left")
hmList_PWM_OKI$introns   <- feature_coverage(current_GB, intronicParts(annots$Oki),                 win = 2000, lab = "Oki introns", direction = "left")
hmList_PWM_OKI$cds       <- feature_coverage(current_GB, cds(annots$Oki),                           win = 2000, lab = "Oki cds", direction = "left")
hmList_PWM_OKI$utrs5     <- feature_coverage(current_GB, unlist(fiveUTRsByTranscript(annots$Oki)),  win = 2000, lab = "Oki 5′ UTRs", direction = "left")
hmList_PWM_OKI$utrs3     <- feature_coverage(current_GB, unlist(threeUTRsByTranscript(annots$Oki)), win = 2000, lab = "Oki 3′ UTRs", direction = "left")

plotHeatmapMeta(hmList_PWM_OKI)
plotHeatmapMeta(hmList_PWM_OKI[c("exons", "introns")])
plotHeatmapMeta(hmList_PWM_OKI[c("exons", "utrs5", "utrs3")])
plotHeatmapMeta(hmList_PWM_OKI[c("utrs5", "utrs3")])
```

They are enriched in introns...

```{r genomic_features3, cache=FALSE}
# 
# # Sandbox to make sanity checks.
# 
# a <- coa$Oki_Osa
# b <- unlist(fiveUTRsByTranscript(annots$Oki))
# 
# hm_a <- feature_coverage(a, b, win = 2000, lab = "All ranges", direction = "left")
# hm_b <- feature_coverage(subsetByOverlaps(a,  b), b, win = 2000, lab = "Subsetted ranges", direction = "left")
# 
# plotHeatmapMeta(list(hm_a))
# plotHeatmapMeta(list(hm_b))
```


## Okinawa genome coverage

As previously discussed, low coverage over an alignment stop could lower the likelihood of it being considered as a breakpoint. We have per-base coverage depth information for the Okinawan genome (`Oki_cov_pb`). However, the coverage of this particular assembly is quite good. In fact, we can investigate the coverge over alignment stops from the information obtained using `master_bp_analysis`. 

```{r echo=T, message=F, warning=F}
# fin_Oki <- fin_gr_O_Oki[[2]]
# min_cov <- min(min(fin_Oki$left_cov_pb), min(fin_Oki$right_cov_pb)) # minimum coverage over an alignment stop
# min_cov
# length(fin_Oki[fin_Oki$left_cov_pb == min_cov]) + length(fin_Oki[fin_Oki$right_cov_pb == min_cov]) # how many of the minimum coverage is observed
# length(fin_Oki[fin_Oki$left_cov_pb <= 50]) + length(fin_Oki[fin_Oki$right_cov_pb <= 50]) # how many alignment stops have a coverage of less than or euqal to 50

```

We may choose to kick out the one alignment stop for which there is no coverage. However, only 15 out of 34572 alignment stops have a coverage of less than or equal to 50, meaning that it would be hard to exclude more than just a few alignment stops using coverage information for breakpoint analysis. 



