---
title: "Scaffolding of the KUM-M3 genome"
author: 
 - "Charles Plessy"
 - "Michael Mansfield"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Scaffolding of the KUM-M3 genome}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(cache = TRUE,
                      dev    = c('png', 'svg'),
                      fig.ext= c('png', 'svg'),
                      fig.width  = 10,
                      fig.height = 10)
```

# Introduction

This vignette produces a scaffolding guide to be used with the KUM-M3 genome for
plotting purposes.

## Load R pacakges and data

```{r load_packages_and_data}
library('OikScrambling') |> suppressPackageStartupMessages()
library('patchwork')     |> suppressPackageStartupMessages()
ggplot2::theme_set(theme_bw())
genomes <- OikScrambling:::loadAllGenomes()
load("BreakPoints.Rdata")
```

# Scaffolding

## Helper functions

```{r Oki_Kum_sc_helpers}
# Helper function to flip many contigs at once.
flipContigsForOxfordPlot <- function(gb, flip_which=NULL) {
  grl <- split(gb, seqnames(gb$query))
  grl <- endoapply(grl, function(s) {
    if( ! identical(unique(as.character(seqnames(s$query))), character(0)) ) {
      if(unique(seqnames(s$query)) %in% flip_which ) {
        s <- GenomicBreaks::reverse(s, query=TRUE)
        s
      }
    }
    s
  })
  ogb <- unlist(grl) |> unname()
  ogb
}

# Let's remove the contigs that have their main match elsewhere.
QTcoverage <- function(gb) {
  stopifnot (length(seqlevels(gb)) == 1) # Not ready for full objects
  stopifnot (!any(is.na(seqlengths(gb$query))))
  grl <- split(gb, seqnames(gb$query))
  lapply(grl, \(gb) sum(width(gb$query))) |> unlist() / seqlengths(gb$query)
}
```


# Scaffolding of Kume chromosomes

The Kume draft assembly contains long contigs that can be scaffolded into
chromosomes.  Some sequences may be missing between these contigs, but the
absence of large gaps in the plots at the junction between contigs suggest that
this is not a problem for the use we make of this scaffolding.

## Chromosome 1

The best match for chr1 in the Kume assembly is `contig_3_1`.  It maps roughly
in the 8,000,000 – 13,000,000 region.

```{r Kum_chr1_assembly_1}
bestMatch <- function(gb)
  tapply(width(gb$query), seqnames(gb$query), sum) |> sort() |> tail(1) |> names()

coa$Oki_Kum |> plyranges::filter(seqnames == "chr1") |> bestMatch()
```

The next best match, `contig_27_1` is most of chr1's short arm.

```{r Kum_chr1_assembly_2}
coa$Oki_Kum |> plyranges::filter(seqnames == "chr1", seqnames(query)!="contig_3_1") |> bestMatch()
coa$Oki_Kum |> plotApairOfChrs(chrQ = "contig_27_1")
```

```{r Kum_chr1_assembly_3}
Oki_Kum_Chr1 <- gbs$Oki_Kum |> plyranges::filter(seqnames == "chr1")
seqlevels(Oki_Kum_Chr1)       <- seqlevelsInUse(Oki_Kum_Chr1)
seqlevels(Oki_Kum_Chr1$query) <- seqlevelsInUse(Oki_Kum_Chr1$query)
makeOxfordPlots(Oki_Kum_Chr1, col='strand') + ggtitle("Everything from KUM-M3 that matches Oki's Chr1")
```

```{r Kum_chr1_assembly_4}
scafs$Kum_Oki <- list(Chr1 = data.frame(
  contig = c("contig_27_1", "contig_90_1", "contig_3_1", "contig_88_1", "contig_87_1", "contig_63_1"),
  orientation = c(1, -1, 1, -1, 1, 1)
))

contigsToMerge_start  <- c(1, cumsum(seqlengths(gbs$Oki_Kum$query)[scafs$Kum_Oki$Chr1$contig]))    |> unname() |> head(-1)
contigsToMerge_end    <-     (cumsum(seqlengths(coa$Oki_Kum$query)[scafs$Kum_Oki$Chr1$contig]) -1) |> unname()
contigsToMerge_oldPos <- GRanges("chr1", IRanges(contigsToMerge_start, contigsToMerge_end))
names(contigsToMerge_oldPos)  <- scafs$Kum_Oki$Chr1$contig
contigsToMerge_oldPos
annot <- genoPlotR::annotation(x1=start(contigsToMerge_oldPos), x2=end(contigsToMerge_oldPos), text=names(contigsToMerge_oldPos), rot=30)

gbs$Oki_Kum |> swap() |> scaffoldByFlipAndMerge(scafs$Kum_Oki) |> plotApairOfChrs(dna_seg_scale=TRUE, annotations = annot)
```

## Chromosome 2

```{r Oki_Kum_sc_chr2_1}
Oki_Kum_Chr2 <- gbs$Oki_Kum |> plyranges::filter(seqnames == "chr2")
seqlevels(Oki_Kum_Chr2)       <- seqlevelsInUse(Oki_Kum_Chr2)
seqlevels(Oki_Kum_Chr2$query) <- seqlevelsInUse(Oki_Kum_Chr2$query)

makeOxfordPlots(Oki_Kum_Chr2, col='strand') + ggtitle("Everything from KUM-M3 that matches Oki's Chr2")
```

Let's remove the contigs that have their main match elsewhere.

```{r Oki_Kum_sc_chr2_2}
Oki_Kum_Chr2[seqnames(Oki_Kum_Chr2$query) %in% seqlevels(Oki_Kum_Chr2$query)[QTcoverage(Oki_Kum_Chr2) < 0.5]] <- NULL
seqlevels(Oki_Kum_Chr2$query) <- seqlevelsInUse(Oki_Kum_Chr2$query)
makeOxfordPlots(Oki_Kum_Chr2, col='strand') + ggtitle("Main hits from KUM-M3 to OKI's Chr2")
```

```{r Kum_chr2_contigs_to_merge}
scafs$Kum_Oki[["Chr2"]] <- data.frame(
  contig = c("contig_36_1", "contig_16_1", "contig_24_1", "contig_79_1", "contig_37_1", "contig_81_1"),
  orientation = c(-1, 1, 1, -1, -1, -1)
)

contigsToMerge_start  <- c(1, cumsum(seqlengths(gbs$Oki_Kum$query)[scafs$Kum_Oki$Chr2$contig]))    |> unname() |> head(-1)
contigsToMerge_end    <-     (cumsum(seqlengths(coa$Oki_Kum$query)[scafs$Kum_Oki$Chr2$contig]) -1) |> unname()
contigsToMerge_oldPos <- GRanges("chr2", IRanges(contigsToMerge_start, contigsToMerge_end))
names(contigsToMerge_oldPos)  <- scafs$Kum_Oki$Chr2$contig
contigsToMerge_oldPos
annot <- genoPlotR::annotation(x1=start(contigsToMerge_oldPos), x2=end(contigsToMerge_oldPos), text=names(contigsToMerge_oldPos), rot=30)

gbs$Oki_Kum |> swap() |> scaffoldByFlipAndMerge(scafs$Kum_Oki) |> plotApairOfChrs(chrT = 'Chr2', dna_seg_scale=TRUE, annotations = annot)
```

## PAR

```{r Oki_Kum_sc_PAR_1}
Oki_Kum_PAR <- gbs$Oki_Kum |> plyranges::filter(seqnames == "PAR")
seqlevels(Oki_Kum_PAR)       <- seqlevelsInUse(Oki_Kum_PAR)
seqlevels(Oki_Kum_PAR$query) <- seqlevelsInUse(Oki_Kum_PAR$query)

makeOxfordPlots(Oki_Kum_PAR, col='strand') + ggtitle("Everything from KUM-M3 that matches OKI's PAR")
```

Let's remove the contigs that have their main match elsewhere.

```{r Oki_Kum_sc_PAR_2}
Oki_Kum_PAR[seqnames(Oki_Kum_PAR$query) %in% seqlevels(Oki_Kum_PAR$query)[QTcoverage(Oki_Kum_PAR) < 0.5]] <- NULL
seqlevels(Oki_Kum_PAR$query) <- seqlevelsInUse(Oki_Kum_PAR$query)
makeOxfordPlots(Oki_Kum_PAR, col='strand') + ggtitle("Main hits from KUM-M3 to OKI's PAR")
```

Let's flip things around a little to get things as linear as they can be:

```{r Oki_Kum_sc_PAR_3}
makeOxfordPlots( flipContigsForOxfordPlot(Oki_Kum_PAR, flip_which = c('contig_12_1', 'contig_43_1')), col='strand')  + ggtitle("KUM-M3 contigs scaffolded and reoriented to PAR")

```

```{r Kum_PAR_contigs_to_merge}
scafs$Kum_Oki[["PAR"]] <- data.frame(
  contig =  c('contig_13_1', 'contig_12_1', 'contig_41_1', 'contig_43_1', 'contig_82_1', 'contig_89_1', 'contig_85_1', 'contig_28_1', 'contig_91_1'),
  orientation = c(1, -1, 1, -1, 1, 1, 1, 1, 1)
)

contigsToMerge_start  <- c(1, cumsum(seqlengths(gbs$Oki_Kum$query)[scafs$Kum_Oki$PAR$contig]))    |> unname() |> head(-1)
contigsToMerge_end    <-     (cumsum(seqlengths(coa$Oki_Kum$query)[scafs$Kum_Oki$PAR$contig]) -1) |> unname()
contigsToMerge_oldPos <- GRanges("chr2", IRanges(contigsToMerge_start, contigsToMerge_end))
names(contigsToMerge_oldPos)  <- scafs$Kum_Oki$PAR$contig
contigsToMerge_oldPos
annot <- genoPlotR::annotation(x1=start(contigsToMerge_oldPos), x2=end(contigsToMerge_oldPos), text=names(contigsToMerge_oldPos), rot=30)

gbs$Oki_Kum |> swap() |> scaffoldByFlipAndMerge(scafs$Kum_Oki) |> plotApairOfChrs(chrT = 'PAR', dna_seg_scale=TRUE, annotations = annot)
```

## XSR

XSR is already complete.

```{r Oki_Kum_sc_XSR_1}
Oki_Kum_XSR <- gbs$Oki_Kum |> plyranges::filter(seqnames == "XSR")
seqlevels(Oki_Kum_XSR)       <- seqlevelsInUse(Oki_Kum_XSR)
seqlevels(Oki_Kum_XSR$query) <- seqlevelsInUse(Oki_Kum_XSR$query)

makeOxfordPlots(Oki_Kum_XSR, col='strand') + ggtitle("Everything from KUM-M3 that matches OKI's XSR")

scafs$Kum_Oki[["XSR"]] <- data.frame(
  contig =  c('contig_42_1'),
  orientation = c(1)
)
```

## YSR

```{r Oki_Kum_sc_ysr}
Oki_Kum_YSR <- gbs$Oki_Kum |> plyranges::filter(seqnames == "YSR")
seqlevels(Oki_Kum_YSR)       <- seqlevelsInUse(Oki_Kum_YSR)
seqlevels(Oki_Kum_YSR$query) <- seqlevelsInUse(Oki_Kum_YSR$query)

makeOxfordPlots(Oki_Kum_YSR, col='strand')

scafs$Kum_Oki[["YSR"]] <- data.frame(
  contig =  c('contig_29_1'),
  orientation = c(-1)
)
```

## Bringing it together

```{r Oki_Kum_sc_all_1}
# Store the results in the package for re-use in other vignettes
# usethis::use_data(scafs, overwrite = TRUE)
identical(scafs$Kum_Oki, OikScrambling::scafs$Kum_Oki)

scaf_Kum_Oki <- scaffoldByFlipAndMerge(gbs$Oki_Kum |> swap(), scafs$Kum_Oki, drop = TRUE) |> swap()
scaf_Kum_Oki |> makeOxfordPlots()

# Original plot for comparison
gbs$Oki_Kum |> makeOxfordPlots()
```
