---
title: "Region of interest: PAC3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Region of interest: PAC3}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{css, echo=FALSE}
pre, code {white-space:pre !important; overflow-x:auto}
```

```{r setup}
knitr::opts_knit$set(cache = TRUE)
options(width = 200)
```

Load packages and data
======================

See `?OikScrambling:::loadAllGenomes()`, `?OikScrambling:::loadAllTranscriptsGR()`,
and `vignette("LoadGenomicBreaks", package = "OikScrambling")` for how the
different objects are prepared.

```{r load_packages_and_data}
library('OikScrambling')   |> suppressPackageStartupMessages()
genomes <- OikScrambling:::loadAllGenomes(compat = F)
load("BreakPoints.Rdata")
```

# PAC3: a gene that moved between chr1 and XSR (but ancestral state is unknown)

## PAC3

The gene g3146 (`Oidioi.mRNA.OKI2018_I69.chr1.g3144.t1.cds` or `CAG5107096` in
OKI2018, `CBY24606` in OdB3) is on `chr1` in Oki and Kum but on `XSR` in Osa and
Bar.  The OdB3 protein contains a non-specific hit for the pfam10178 (Proteasome
assembly chaperone 3) PAC3 domain, and Psi-BLAST also discovers it.

<details>
<summary>Click here to see a provisional alignment with similar sequences in
other larvacean genomes.  Proper attention to splicing would be needed to make
it more accurate.</summary>
```
O_dio       MHPTTILREIELENEPVALSIMKFAEETLVTISDNGTFGAFHDI-DIA-EMRPGKEPLIT
O_dio_Oska  MHPETILREVEIDNSPVALAIVKFADQTMVTISDNGTFGAFHDI-EIA-EMRPGTDPVVT
O_van       ----SIVESTQIEGVDVSLAIMGFADCTMIAVSDTGTFGSFYKV-QVAKGGHTNQEAVVT
O_alb       ---------LDLENADIHLAAMSFSDCLMLAVSDTGTFGSFYQV-EKSETSHQNIEPVIS
O_lon       ---------IYLGEEKVHIAVMGFGDATMITVSDGALFGPIYKV-TKS--QKGANEPVIT
O_lon_N     --------EIEIDNEVCYISVTQFGDTNFVTISDTGTFGSFYQV-EKK--EKMGNEPIFN
B_sty       ---------FEIDDLVIYFSFMEFDDYVFLALSNTGTFGSFYKV-QKN--FRMDSNPIIS
M_ery       ------IRIIVIDNTEIYLSVMKFGDCNFIVISDTGTFGSFFKI-EKN--FKMDNNPVVN
O_ruf_Y     ------LKRLRLMSQKVHLSSMKFDDHIMLGVSDSGTFWLILSCFFKSTNQHKDKEPVFE


O_dio       IKPF-FGYSDDFTKV------------VCRNLATSLQVKNK-LMISIGLKQEKVNKQNLD
O_dio_Oska  VKPF-FGYSDDFAKVV-----------VCRNLATGLEVKNK-LMISIGLKQERINKNN--
O_van       IDPI-FGINDHYFQVLIR------------------------------------------
O_alb       IRPI-FGVDDQYFQVFLQVLPDSYFKVVCRYLGVHLAESRR-LLVSITLQKDKINKDN--
O_lon       VDGV-FGEQHDYAQVV------------CRHLRLHCRVETEQLLASVLLKRDLVTREN--
O_lon_N     IQPI-FGQAEDYLQVCKFQIETS--IVVCRYLCQNCTLQGPM-MVSVILSREKVNKKN--
B_sty       TEQI-FGNKKDLLEVINK------------------------------------------
M_ery       IEPI-FGENKDYLQV---------------------------------------------
O_ruf_Y     IVPIFFGCDDQYFQV---------------------------------------------
```
</details>

Private ZENBU link: <https://fantom.gsc.riken.jp/zenbu/gLyphs/#config=nBuUjj3vMBkuTVQnE57TC;loc=OKI2018_I69_1.0::chr1:11497331..11498497+>

#### PAC3 in five genome pairs

```{r PAC3}
PAC3 <- GRanges("chr1:11497331-11498497")
PAC3_in_Kum <- GRanges("contig_3_1:3247998-3248575")

coa$Oki_Osa |> subsetByOverlaps(PAC3)
coa$Oki_Aom |> subsetByOverlaps(PAC3)
coa$Oki_Bar |> subsetByOverlaps(PAC3)
coa$Oki_Nor |> subsetByOverlaps(PAC3)
coa$Oki_Kum |> subsetByOverlaps(PAC3)
```

## The bridge region of PAC3 in the "Northern" genomes.

```{r PAC3_bridge}
PAC3_bridge_Osa <- coa$Oki_Osa |> subsetByOverlaps(PAC3) |> swap() |> cleanGaps() |> plyranges::mutate(strand = '-')
PAC3_bridge_Aom <- coa$Oki_Aom |> subsetByOverlaps(PAC3) |> swap() |> cleanGaps() |> plyranges::mutate(strand = '-')
PAC3_bridge_Bar <- coa$Oki_Bar |> subsetByOverlaps(PAC3) |> swap() |> cleanGaps()
PAC3_bridge_Nor <- coa$Oki_Nor |> subsetByOverlaps(PAC3) |> swap() |> cleanGaps()

list( Osa = PAC3_bridge_Osa
    , Aom = PAC3_bridge_Aom
    , Bar = PAC3_bridge_Bar
    , Nor = PAC3_bridge_Nor) |> lapply(\(gb) gb + 0)  |> lapply(getSeq) |> lapply(unlist) |> as("DNAStringSet") |> msa::msaClustalW() |> as("DNAMultipleAlignment")
```

## The left region of PAC3 in all genomes

```{r PAC3_left}
PAC3_left_Oki_Osa  <- coa$Oki_Osa |> subsetByOverlaps(PAC3) |> plyranges::slice(1) |> get_bps("right")
PAC3_left_Kum_Osa  <- coa$Osa_Kum |> swap () |> subsetByOverlaps(PAC3_in_Kum + 1) |> sort(i=T) |> plyranges::slice(1) |> get_bps("right")
PAC3_left_Osa      <- coa$Oki_Osa |> subsetByOverlaps(PAC3) |> swap() |> plyranges::slice(1) |> get_bps("left") |> plyranges::mutate(strand = '-')
PAC3_left_Aom      <- coa$Oki_Aom |> subsetByOverlaps(PAC3) |> swap() |> plyranges::slice(1) |> get_bps("left") |> plyranges::mutate(strand = '-')
PAC3_left_Oki_Bar  <- coa$Oki_Bar |> subsetByOverlaps(PAC3) |> plyranges::slice(1) |> get_bps("right")
PAC3_left_Bar      <- coa$Oki_Bar |> subsetByOverlaps(PAC3) |> swap()  |> plyranges::slice(1) |> get_bps("right")
PAC3_left_Nor      <- coa$Oki_Nor |> subsetByOverlaps(PAC3) |> swap()  |> plyranges::slice(1) |> get_bps("right")

# Shifting Bar and Nor so that they all align well

PAC3_left <- 
list( Aom = PAC3_left_Aom
    , Osa = PAC3_left_Osa
    , Oki = PAC3_left_Oki_Osa
    , Kum = PAC3_left_Kum_Osa
    , Bar = PAC3_left_Bar |> shift(20)
    , Nor = PAC3_left_Nor |> shift(20))

PAC3_left |> lapply(\(gb) gb + 70) |> lapply(getSeq) |> lapply(unlist) |> as("DNAStringSet") |> msa::msaClustalW() |> as("DNAMultipleAlignment")
```

## The right region of PAC3 in all genomes

Very near the alignment break (at the center), there is the `ATG` and upstream
of it there are two conserved `AG` that may be trans-splicing sites
(with CAGE support at least in Oki).

```{r PAC3_right}
PAC3_right_Oki_Osa  <- coa$Oki_Osa |> subsetByOverlaps(PAC3) |> plyranges::slice(3) |> get_bps("left")
PAC3_right_Kum_Osa  <- coa$Osa_Kum |> swap () |> subsetByOverlaps(PAC3_in_Kum + 1) |> sort(i=T) |> plyranges::slice(3) |> get_bps("left")
PAC3_right_Osa      <- coa$Oki_Osa |> subsetByOverlaps(PAC3) |> swap() |> plyranges::slice(3) |> get_bps("right") |> plyranges::mutate(strand = '-')
PAC3_right_Aom      <- coa$Oki_Aom |> subsetByOverlaps(PAC3) |> swap() |> plyranges::slice(3) |> get_bps("right") |> plyranges::mutate(strand = '-')
PAC3_right_Oki_Bar  <- coa$Oki_Bar |> subsetByOverlaps(PAC3) |> plyranges::slice(3) |> get_bps("left")
PAC3_right_Bar      <- coa$Oki_Bar |> subsetByOverlaps(PAC3) |> swap()  |> plyranges::slice(3) |> get_bps("left")
PAC3_right_Nor      <- coa$Oki_Nor |> subsetByOverlaps(PAC3) |> swap()  |> plyranges::slice(3) |> get_bps("left")

PAC3_right <-
list( Aom = PAC3_right_Aom
    , Osa = PAC3_right_Osa
    , Oki = PAC3_right_Oki_Osa
    , Kum = PAC3_right_Kum_Osa
    , Bar = PAC3_right_Bar
    , Nor = PAC3_right_Nor)
PAC3_right |> lapply(\(gb) gb + 70) |> lapply(getSeq) |> lapply(unlist) |> as("DNAStringSet") |> msa::msaClustalW() |> as("DNAMultipleAlignment") 
```

## Combined view of the left and right regions

```{r PAC3_left_right}
PAC3_left_right <- list(
  Aom  = range(PAC3_left_Aom, PAC3_right_Aom),
  Osa  = range(PAC3_left_Osa, PAC3_right_Osa),
  OkiL = PAC3_left_Oki_Osa,
  OkiR = PAC3_right_Oki_Osa,
  KumL = PAC3_left_Kum_Osa,
  KumR = PAC3_right_Kum_Osa,
  Bar  = range(PAC3_left_Bar |> shift(20), PAC3_right_Bar),
  Nor  = range(PAC3_left_Nor |> shift(20), PAC3_right_Nor)
)
PAC3_left_right.aln <- PAC3_left_right |>
  lapply(\(gb) gb + 50) |>
  lapply(getSeq) |>
  lapply(unlist) |>
  as("DNAStringSet") |>
  msa::msaClustalOmega() |>
  as("DNAMultipleAlignment")

# Re-order based on my preference
PAC3_left_right.aln@unmasked[c(1,2,5,6,7,8,3,4)]
```

```{r PAC3_XSR}
PAC3_XSR_Osa    <- coa$Oki_Osa |> subsetByOverlaps(PAC3) |> plyranges::slice(2) |> swap() 
PAC3_XSR_Aom    <- coa$Oki_Aom |> subsetByOverlaps(PAC3) |> plyranges::slice(2) |> swap()
PAC3_XSR_Bar    <- coa$Oki_Bar |> subsetByOverlaps(PAC3) |> plyranges::slice(2) |> swap()
PAC3_XSR_Nor    <- coa$Oki_Nor |> subsetByOverlaps(PAC3) |> plyranges::slice(2) |> swap()

list( Osa = PAC3_XSR_Osa
    , Aom = PAC3_XSR_Aom
    , Bar = PAC3_XSR_Bar
    , Nor = PAC3_XSR_Nor) |> lapply(getSeq) |> lapply(unlist) |> as("DNAStringSet") |> msa::msaMuscle() |> as("DNAMultipleAlignment") |> msaR::msaR()
```

# Session information

```{r session_information}
sessionInfo()
```
