---
title: "Translocations"
author: 
 - "Charles Plessy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Translocations}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE)
knitr::opts_knit$set(verbose = TRUE)
```

# Introduction

After coalescing colinear alignments, removing translocations of
repeat-containing sequences and re-coalescing, colinearity is still broken
hundreds of time.

Here we explore the role of translocations in scrambling _Oikopleura_ genomes.

# Load R pacakges and data

```{r load_packages_and_data}
library('OikScrambling') |> suppressPackageStartupMessages()
(genomes <- OikScrambling:::loadAllGenomes())
(annots <- OikScrambling:::loadAllAnnotations() |> suppressWarnings())
(reps <- OikScrambling:::loadAllRepeats())
(load("BreakPoints.Rdata"))
```

See `vignette("LoadGenomicBreaks", package = "OikScrambling")` for how the
different GBreaks objects are prepared.

# Translocations

## Documentation

Details can be found in `vignette("GenomicBreaks", package = "GenomicBreaks")`,
in `vignette("StructuralVariants", package = "GenomicBreaks")`,
and `?GenomicBreaks::flagTranslocations`

## Number of translocations

Notes:

 - Translocations in a genomic context ressembling insertion/deletion are removed
   in the `coa2` objects.
   
 - At the moment, the `flagTranslocations` function only searches the _target_
   genome.

```{r detect_trivial_invertions}
(trans_summary <- data.frame(
  align  = sapply(gbs,  function(gb) sum(flagTranslocations(gb)$tra)),
  mapped = sapply(coa,  function(gb) sum(flagTranslocations(gb)$tra)),
  map2   = sapply(coa2, function(gb) sum(flagTranslocations(gb)$tra))
))

trans_summary$pairname  <- OikScrambling:::compDistance(rownames(trans_summary))
trans_summary$target    <- sub("_.*", "", rownames(trans_summary))

sapply(trans_summary[,c("align", "mapped", "map2")], \(x) tapply(x, trans_summary$pairname, mean))
sapply(trans_summary[,c("align", "mapped", "map2")], \(x) tapply(x, trans_summary$pairname, median))
sapply(trans_summary[,c("align", "mapped", "map2")], \(x) tapply(x, trans_summary$pairname, sd))

ggplot(trans_summary |> tidyr::pivot_longer(c("align", "mapped", "map2"))) +
  aes(value, name) + geom_point(aes(col = pairname, pch = target)) +
  ggtitle ("Number of translocations found after different post-processings")
```
